---
title: "phackR rounding"
author: "Carolin Kirsch"
date: "2026-07-20"
output: html_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(phackR)
library(dplyr)
library(tidyr)
library(scales)
library(flextable)
library(apa7)

set_flextable_defaults(theme_fun = theme_apa,
                        font.family = "Times New Roman")
```

# incorrectRounding.R

Lines below are copied from the `phackR` package (astefan1/phacking_compendium, MIT License © 2019 Angelika Stefan):
<https://github.com/astefan1/phacking_compendium/blob/master/phackR/R/incorrectRounding.R>
```{r eval=FALSE}
# ==============================================================================
# Incorrect Rounding
# ==============================================================================

# Generic sampling function .sim.data() can be used

#' P-Hacking function for incorrect rounding
#' @description Outputs a p-hacked p-value and the non-p-hacked-p-value
#' @param df Data frame
#' @param group Scalar defining location of the group vector in the data frame
#' @param dv Scalar defining location of dependent variable in the data frame
#' @param roundinglevel Highest p-value that is rounded down to 0.05
#' @param alternative Direction of the t-test ("two.sided", "less", "greater")
#' @param alpha Significance level of the t-test (default: 0.05)
#' @importFrom stats t.test

.roundhack <- function(df, group, dv, roundinglevel, alternative = "two.sided", alpha = 0.05){

  # Compute t-test
  pval <- stats::t.test(df[,dv] ~ df[,group],
                        var.equal = TRUE, alternative = alternative)$p.value
  r2val <- .compR2t(df[,dv][(df[,group] == unique(df[,group])[1])],
                    df[,dv][(df[,group] == unique(df[,group])[2])])

  # P-hack p-value
  if(pval > alpha && pval < roundinglevel){
    p.final <- alpha
  } else {
    p.final <- pval
  }

  ps <- c(pval, p.final)

  return(list(p.final = p.final,
              ps = ps,
              r2.final = r2val,
              r2s = rep(r2val, 2)))

}

#' Simulate p-hacking with incorrect rounding
#' @param roundinglevel Highest p-value that is rounded down to alpha
#' @param iter Number of iterations
#' @param alternative Direction of the t-test ("two.sided", "less", "greater")
#' @param alpha Significance level of the t-test (default: 0.05)
#' @param shinyEnv Is the function run in a Shiny session? TRUE/FALSE
#' @export

sim.roundhack <- function(roundinglevel, iter = 1000, alternative = "two.sided", alpha = 0.05, shinyEnv = FALSE){

  # Simulate as many datasets as desired iterations
  dat <- list()
  for(i in 1:iter){
    dat[[i]] <- .sim.data(nobs.group = 30)
  }

  # Apply p-hacking procedure to each dataset
  if(!shinyEnv){
    res <- pbapply::pblapply(dat, .roundhack, group = 1, dv = 2,
                  roundinglevel = roundinglevel, alternative = alternative, alpha = alpha)
  }
  if(shinyEnv){
    percentage <- 0
    withProgress(message = "Running simulation", value = 0, {
      res = lapply(dat, function(x){
        percentage <<- percentage + 1/length(dat)*100
        incProgress(1/length(dat), detail = paste0("Progress: ",round(percentage,2), "%"))
        .roundhack(df=x, group = 1, dv = 2, roundinglevel = roundinglevel,
                   alternative = alternative, alpha = alpha)
      })
    })
  }

  ps.hack <- NULL
  ps.orig <- NULL
  r2s.hack <- NULL
  r2s.orig <- NULL

  for(i in 1:iter){
    ps.hack[i] <- res[[i]][["p.final"]]
    ps.orig[i] <- res[[i]][["ps"]][1]
    r2s.hack[i] <- res[[i]][["r2.final"]]
    r2s.orig[i] <- res[[i]][["r2s"]][1]
  }

  res <- cbind(ps.hack, ps.orig, r2s.hack, r2s.orig)

  return(res)

}
```
hard-coded n=30

# Simulation

Lines below are copied from the original `roundHack_simulation.R` (astefan1/phacking_compendium, MIT License © 2019 Angelika Stefan), with a fixed seed and caching added:
<https://github.com/astefan1/phacking_compendium/blob/master/simulations/roundHack_simulation.R>
```{r roundHack_simulation}
# Incorrect Rounding: Simulation

#### Conditions ####
roundinglevel <- c(0.051, 0.055, 0.1)

cond.roundHack <- expand.grid(roundinglevel)

#### Simulation ####

set.seed(1495)

simmultiple.roundHack <- function(par, strategy){
  data.frame(sim.roundhack(roundinglevel = par[1],
                         iter = 10000,
                         alpha = 0.05))
  
  
}

roundHack_cache_file <- "simresults_roundHack_repro.rds"

if (file.exists(roundHack_cache_file)) {
  simresults.roundHack <- readRDS(roundHack_cache_file)
} else {
  simresults.roundHack <- apply(cond.roundHack, 1, function(x) {
    simmultiple.roundHack(x)
  })

  saveRDS(simresults.roundHack, roundHack_cache_file)
}
```


# Results
## reproduced plot

The relevant part below is copied from the original `plots_FPrate.R` (astefan1/phacking_compendium, MIT License © 2019 Angelika Stefan):
<https://github.com/astefan1/phacking_compendium/blob/master/simulations/plots_FPrate.R>
```{r roudHack_simulation}
#### Incorrect Rounding ####

roundinglevel <- c(0.051, 0.055, 0.1)
cond.roundHack <- expand.grid(roundinglevel)

FP.rates <- sapply(simresults.roundHack,
       function(x) {sum(x$ps.hack < 0.05) / nrow(x)})

cond.roundHack$FP.rates <- FP.rates

ggplot(cond.roundHack, aes(Var1, FP.rates)) +
  geom_point(size = 2) +
  geom_line() +
  geom_hline(yintercept = 0.05, col = "grey") +
  theme_classic() +
  scale_y_continuous(breaks = c(0, 0.05, 0.1, 0.2, 0.3, 0.4), limits = c(0,0.45)) +
  labs(title = "False Positive Rates",
       x = "Rounding Level",
       y = "False Positive Rate")

# This plot is actually misleading because the FP rates are dependent on the 
# sampling variability, i.e., if we want to be able to get a proper estimate
# of the small differences in FP rates, we would have to conduct many more
# simulations.

```

## fixed FPR

```{r}
sim_summary_roundHack_fixed <- data.frame(
  n = 30,
  p_max = cond.roundHack$Var1
) |>
  mutate(
    empirical_detection_rate = sapply(simresults.roundHack, function(x) mean(x$ps.hack <= 0.05)), #inclusive rejection criterion!
    empirical_detection_rate_mcse = sqrt(empirical_detection_rate * (1 - empirical_detection_rate) / sapply(simresults.roundHack, nrow))
  )

sim_summary_roundHack_fixed
```

## merged table

```{r}
merged_table <- data.frame(
  p_max = cond.roundHack$Var1,
  `reproduced` = scales::number(cond.roundHack$FP.rates, accuracy = 0.001),
  `fixed`= paste0(scales::number(sim_summary_roundHack_fixed$empirical_detection_rate, accuracy = 0.001),
                       " (",
                       scales::number(sim_summary_roundHack_fixed$empirical_detection_rate_mcse, accuracy = 0.001),
                       ")")
)

merged_table |>
  mutate(p_max = as.character(p_max)) |>
  rename(`rounding threshold` = p_max) |>
  column_spanner_label("estimated FPR", -1) |>
  apa_flextable(no_format_columns = `rounding threshold`) |>
  align(j = 1, align = "center")
```

