---
title: "alternative hypothesis tests"
author: "Carolin Kirsch"
date: "2025-12-27"
output: html_document
---

```{r}
library(dplyr)
library(tidyverse)
library(tidyr)
library(ggplot2)
library(ggtext)
library(kableExtra)
library(DescTools)
```

```{r}
source("functions.R")
```

```{r}
n_iterations <- 10000
alpha <- 0.05
```


Use alternative hypothesis tests to achieve statistical significance.Conducting an independent-samples t-test, a Welch test, a Wilcoxon test and a Yuen test with four different trimming parameters,c ∈ {0.1, 0.15, 0.2, 0.25}, on the same dataset.

# functions
## data generating
source generate_data_ttest()

## data analysis / p-hacking
```{r}
analyse_data_alternative_tests <- function(data) {
  tibble(
    test = c("students", "welch", "wilcox", 
             "yuen0.1", "yuen0.15", "yuen0.2", "yuen0.25"),
    p_value = c(
      t.test(score ~ condition, data, alternative = "two.sided", var.equal = TRUE)$p.value,
      t.test(score ~ condition, data, alternative = "two.sided", var.equal = FALSE)$p.value,
      wilcox.test(score ~ condition, data, alternative = "two.sided")$p.value,
      YuenTTest(score ~ condition, data, alternative = "two.sided", trim = 0.1)$p.value,
      YuenTTest(score ~ condition, data, alternative = "two.sided", trim = 0.15)$p.value,
      YuenTTest(score ~ condition, data, alternative = "two.sided", trim = 0.2)$p.value,
      YuenTTest(score ~ condition, data, alternative = "two.sided", trim = 0.25)$p.value
    )
  )
}
```

TO DO: estimate FPR per test -> should be .05, combination should be inflated -> multiple tests

## reporting strategy
sourcing reporting_strategy_tibbles()

# simulation
```{r}
experiment_parameters_grid <- expand_grid(
  n_per_group = c(30, 50, 100, 300),
  iteration = 1:n_iterations,
  reporting_strategy = c("first significant","smallest", "smallest significant")
)
```

```{r}
set.seed(12345)

simulation <- experiment_parameters_grid |>
  mutate(
    generated_data = pmap(
      list(n_per_group, n_per_group),
      generate_data_ttest
    ),
    analysis_result = map(
      generated_data,
      analyse_data_alternative_tests
    ),
    reported_result = pmap(
      list(analysis_result, reporting_strategy),
      reporting_strategy_tibbles
    )
  )

saveRDS(simulation, "results/sim_alternative_tests.rds")
```

# results
## table
```{r}
# summarise simulation results over the iterations ----
simulation_summary <- simulation |>
  unnest(reported_result) |> 
  group_by(n_per_group) |>
  summarize(FPR = janitor::round_half_up(mean(p_value < .05), 2))

save(simulation_summary, "results/sim_summary_alterantive_tests.rds")

simulation_summary
```

## plot

### neded for replication

```{r}
ggplot(simulation_summary, 
       aes(x=n_per_group,
           y=FPR)) +
  geom_line() +
  geom_point() +
  labs(
    x = "sample size",
    y = "false positive rate") +
  ylim(0,0.5) +
  scale_x_continuous(breaks = c(30, 50, 100, 300)) +
  geom_hline(yintercept=alpha, colour="grey") +
  theme_classic()
```

### add-ons?

```{r}
which_test_percentage_FPR <- simulation |>
  unnest(reported_result) |>
  filter(p_value <= alpha) |>
  count(test) |>
  mutate(percentage = n / sum(n))

which_test_percentage_FPR
```


```{r}
test_count_FPR_reporting <- simulation |>
  unnest(reported_result) |>
  filter(p_value <= alpha) |>
  count(reporting_strategy, test) |>
  group_by(reporting_strategy) |>
  mutate(percentage_within_reporting_strategy = n / sum(n)) |>
  ungroup()

test_count_FPR_reporting
```


```{r}
which_test_count_smallest <- simulation |>
  filter(reporting_strategy == "smallest") |> 
  unnest(reported_result) |> 
  count(test) |> 
  mutate(percentage_within_smallest = n/sum(n))

which_test_count_smallest
```

```{r}
plot_data <- simulation |>
  unnest(reported_result) |>
  filter(p_value <= alpha) |>
  count(reporting_strategy, test) |>
  group_by(reporting_strategy) |>
  mutate(prop = n / sum(n))

ggplot(plot_data, aes(x = reporting_strategy, y = prop, fill = test)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    x = "Reporting strategy",
    y = "Proportion of reported significant results",
    fill = "Reported test"
  ) +
  theme_minimal()

```

OPEN QUESTION: would it be interesting to see this for different sample sizes as well? resp. check if there is a difference
