---
title: "alternative hypothesis tests"
author: "Carolin Kirsch"
date: "2025-12-27"
output:
  html_document: default
  word_document: default
  pdf_document:
    latex_engine: xelatex
---

```{r setup}
library(dplyr)
library(tidyverse)
library(tidyr)
library(ggplot2)
library(ggtext)
library(flextable)
library(apa7)
library(ftExtra)
library(DescTools)

set_flextable_defaults(theme_fun = theme_apa,
                        font.family = "Times New Roman")
```

```{r source}
source("functions.R")
```

```{r general_seettings}
n_iterations <- 10000L
alpha <- 0.05
```

Use alternative hypothesis tests to achieve statistical significance.Conducting an independent-samples t-test, a Welch test, a Wilcoxon test and a Yuen test with four different trimming parameters,c ∈ {0.1, 0.15, 0.2, 0.25}, on the same dataset.

# functions
## data analysis alternative tests
```{r alternative_tests}
analyse_data_alternative_tests <- function(data) {
  tibble(
    test = c("students", "welch", "wilcox", 
             "yuen0.1", "yuen0.15", "yuen0.2", "yuen0.25"),
    p_value = c(
      t.test(score ~ condition, data, alternative = "two.sided", var.equal = TRUE)$p.value,
      t.test(score ~ condition, data, alternative = "two.sided", var.equal = FALSE)$p.value,
      wilcox.test(score ~ condition, data, alternative = "two.sided")$p.value,
      YuenTTest(score ~ condition, data, alternative = "two.sided", trim = 0.1)$p.value,
      YuenTTest(score ~ condition, data, alternative = "two.sided", trim = 0.15)$p.value,
      YuenTTest(score ~ condition, data, alternative = "two.sided", trim = 0.2)$p.value,
      YuenTTest(score ~ condition, data, alternative = "two.sided", trim = 0.25)$p.value
    )
  )
}
```

# simulation
```{r sim_par_alternative_tests}
par_grad_alternative_tests <- expand_grid(
  n = c(30, 50, 100, 300),
  iteration = 1:n_iterations,
  reporting_strategy = c("first significant","smallest", "smallest significant")
)
```

```{r sim_alt_tests}
sim_alternative_tests_file <- paste0("results/sim_alternative_tests_",n_iterations,"iterations.rds")

if(!file.exists(sim_alternative_tests_file)){
  set.seed(8974)

  sim_alternative_tests <- par_grad_alternative_tests |>
    mutate(
      generated_data = pmap(
        list(n),
        generate_data_ttest
      ),
      analysis_result = map(
        generated_data,
        analyse_data_alternative_tests
      ),
      reported_result = pmap(
        list(analysis_result, reporting_strategy, alpha),
        reporting_strategy_tibbles
      )
    )

  saveRDS(sim_alternative_tests, sim_alternative_tests_file)
} else {
  sim_alternative_tests <- readRDS(sim_alternative_tests_file)
}
```

# results

## replication table
```{r summary_alt_tests}
sim_summary_alternative_tests_file <- paste0("results/sim_summary_alternative_tests_",n_iterations,"iterations.rds")

if(!file.exists(sim_summary_alternative_tests_file)){
  sim_summary_alternative_tests <- sim_alternative_tests |>
    unnest(reported_result) |>
    group_by(n) |>
    summarize(
      empirical_detection_rate  = mean(p_value < alpha),
      empirical_detection_rate_mcse = sqrt(empirical_detection_rate * (1 - empirical_detection_rate) / n()),
      .groups = "drop"
      ) |>
    mutate(
      empirical_detection_rate_95CI_low = empirical_detection_rate - empirical_detection_rate_mcse * 1.96,
      empirical_detection_rate_95CI_upp = empirical_detection_rate + empirical_detection_rate_mcse * 1.96,
    )

  saveRDS(sim_summary_alternative_tests, sim_summary_alternative_tests_file)
} else {
  sim_summary_alternative_tests <- readRDS(sim_summary_alternative_tests_file)
}

sim_summary_alternative_tests
```

TO DO: row names, note, caption, etc.
```{r flextable_repli_alt_tests}
sim_summary_alternative_tests |>
  mutate(`n per condition` = as.character(n)) |>
  mutate(across(where(is.numeric), \(x) scales::number(x, accuracy = 0.001))) |>
  mutate(edr_string = paste0(empirical_detection_rate,
                             " (", 
                             empirical_detection_rate_mcse,
                             ")")) |>
  select(`n per condition`,
         `empirical FPR (±1 MCSE)` = edr_string) |>
  apa_flextable(no_format_columns = `n per condition`) |>
  align(j = 1, align = "center") |>
  add_footer_lines(values = as_paragraph(as_i("Note. "), "Values in parentheses are ±1 Monte Carlo Standard Error."))
```
<!-- caption for quarto integration: "Empirical False Positive Rate (±1 MCSE) by Sample Size" -- will be set via #| tbl-cap in the .qmd chunk -->


## replication plot
```{r plot_repli_alt_tests}
plot_repli_alternative_tests <- ggplot(sim_summary_alternative_tests, 
       aes(x=n,
           y=empirical_detection_rate)) +
  geom_hline(yintercept=alpha, colour="grey") +
  geom_linerange(aes(ymin = empirical_detection_rate - empirical_detection_rate_mcse,
                     ymax = empirical_detection_rate +empirical_detection_rate_mcse)
                , color = "black"
                ) +
  geom_line() +
  geom_point() +
  labs(
    x = "sample size per group",
    y = "empirical false-positive-rate") +
  theme(text = element_text(size=20)) +
  scale_y_continuous(breaks = c(0, 0.05, 0.1, 0.2, 0.3, 0.4), limits = c(0,0.45)) +
  scale_x_continuous(breaks = c(30, 50, 100, 300)) +
  theme_classic()

plot_repli_alternative_tests
saveRDS(plot_repli_alternative_tests, "results/plot_repli_alternative_tests_10000iterations.rds")
```

## extension 1: FPR per test
```{r summary_FPR_test_n_alt_tests}
sim_summary_test_n_alternative_tests_file <- paste0("results/sim_summary_test_n_alternative_tests_",n_iterations,"iterations.rds")

if(!file.exists(sim_summary_test_n_alternative_tests_file)){
  sim_summary_test_n_alternative_tests <- sim_alternative_tests |>
    unnest(analysis_result) |>
    group_by(n, test) |>
    summarize(
      empirical_detection_rate  = mean(p_value < alpha),
      empirical_detection_rate_mcse = sqrt(empirical_detection_rate * (1 - empirical_detection_rate) / n()),
      .groups = "drop"
      )

  saveRDS(sim_summary_test_n_alternative_tests, sim_summary_test_n_alternative_tests_file)
} else {
  sim_summary_test_n_alternative_tests <- readRDS(sim_summary_test_n_alternative_tests_file)
}

sim_summary_test_n_alternative_tests
```

TO DO: row names, note, caption, etc.
```{r FPR_test_n_flextable_alt_tests}
sim_summary_test_n_alternative_tests |>
  mutate(`n per condition` = as.character(n),
         test = as.character(test)) |>
  mutate(across(where(is.numeric), \(x) scales::number(x, accuracy = 0.001))) |>
  mutate(edr_string = paste0(empirical_detection_rate,
                             " (", 
                             empirical_detection_rate_mcse,
                             ")")) |>
  select(`n per condition`,
         test,
         edr_string) |>
  pivot_wider(names_from = test,
              values_from = edr_string) |>
  column_spanner_label("Hypothesis test", -1) |>
  apa_flextable(no_format_columns = `n per condition`) |>
  align(j = 1, align = "center") |>
  add_footer_lines(values = as_paragraph(as_i("Note. "), "Values in parentheses are ±1 Monte Carlo Standard Error."))
```
<!-- caption for quarto integration: "Empirical False Positive Rate (±1 MCSE) by Hypothesis Test and Sample Size" -- will be set via #| tbl-cap in the .qmd chunk -->


```{r FPR_test_alt_tests}
sim_summary_test_alternative_tests <- sim_alternative_tests |> 
  unnest(analysis_result) |> 
  group_by(test) |>
  summarize(
    empirical_detection_rate  = mean(p_value < alpha),
    empirical_detection_rate_mcse = sqrt(empirical_detection_rate * (1 - empirical_detection_rate) / n()),
    .groups = "drop"
    ) |> 
  mutate(
    empirical_detection_rate_95CI_low = empirical_detection_rate - empirical_detection_rate_mcse * 1.96,
    empirical_detection_rate_95CI_upp = empirical_detection_rate + empirical_detection_rate_mcse * 1.96,
  )

sim_summary_test_alternative_tests
```
OPEN: does this also need a flextable table? or could I maybe include it as sum row in the upper table?

## extension 2: count tests within reported FP results

```{r reported_test_percentage}
all_tests <- c("students", "welch", "wilcox",
               "yuen0.1", "yuen0.15", "yuen0.2", "yuen0.25")

reported_test_percentage <- sim_alternative_tests |>
  unnest(reported_result) |>
  filter(p_value < alpha) |>
  mutate(test = factor(test, levels = all_tests)) |>
  count(test, .drop = FALSE) |>
  mutate(percentage = n / sum(n))

reported_test_percentage
```

```{r investigation_zero_welch}
sim_alternative_tests |> 
  unnest(analysis_result) |> 
  filter(test == "students" | test == "welch") |> 
  pivot_wider(names_from = test, values_from = p_value) |> 
  mutate(p_diff_students_welch = students - welch) |> 
  summarize(min_p_diff_students_welch = min(p_diff_students_welch),
            max_p_diff_students_wekch = max(p_diff_students_welch))
```

```{r test_count_FPR}
altttests_test_per_strategy_file <- paste0("results/altttests_test_per_strategy_",n_iterations,"iterations.rds")

if(!file.exists(altttests_test_per_strategy_file)){
  reported_test_percentage_by_strategy <- sim_alternative_tests |>
    unnest(reported_result) |>
    mutate(test = factor(test, levels = all_tests)) |>
    filter(p_value < alpha) |>
    count(reporting_strategy, test, .drop = FALSE) |>
    group_by(reporting_strategy) |>
    mutate(percentage_within_reporting_strategy = n / sum(n)) |>
    ungroup()

  saveRDS(reported_test_percentage_by_strategy, altttests_test_per_strategy_file)
} else {
  reported_test_percentage_by_strategy <- readRDS(altttests_test_per_strategy_file)
}

reported_test_percentage_by_strategy
```

```{r test_count_FPR_flextable}
reported_test_percentage_by_strategy |>
  mutate(`reporting strategy` = as.character(reporting_strategy),
         test = as.character(test)) |>
  mutate(across(where(is.numeric), \(x) scales::number(x, accuracy = 0.001))) |>
  select(`reporting strategy`,
         test,
         percentage_within_reporting_strategy) |>
  pivot_wider(names_from = test,
              values_from = percentage_within_reporting_strategy) |>
  column_spanner_label("Hypothesis test", -1) |>
  apa_flextable(no_format_columns = `reporting strategy`) |>
  align(j = 1, align = "left") |>
  add_footer_lines(values = as_paragraph(as_i("Note.")))
```
<!-- caption for quarto integration: "Percentages of Type of Hypothesis Test within Reported False-Positive Analysis Results" -- will be set via #| tbl-cap in the .qmd chunk -->


```{r}
ggplot(reported_test_percentage_by_strategy, 
       aes(x = reporting_strategy, 
           y = percentage_within_reporting_strategy, 
           fill = test)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    x = "Reporting strategy",
    y = "Proportion of reported significant results",
    fill = "Reported test"
  ) +
  theme_minimal()
```
OPEN QUESTION: would it be interesting to see this for different sample sizes as well? resp. check if there is a difference

```{r}
sessionInfo()
writeLines(capture.output(sessionInfo()), paste0("results/alttests_sessionInfo_", n_iterations, "iterations.txt"))
```

