---
title: "Discretizing Variables"
author: "Carolin Kirsch"
output:
  html_document:
    code_folding: hide
---

```{r setup, message = FALSE, warning = FALSE, class.source = "fold-show"}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

library(dplyr)
library(tidyverse)
library(tidyr)
library(ggplot2)
library(flextable)
library(apa7)
library(ftExtra)

set_flextable_defaults(theme_fun = theme_apa,
                        font.family = "Times New Roman")
```

```{r source, class.source = "fold-show"}
source("functions.R")
```

```{r general_settings, class.source = "fold-show"}
n_iterations <- 10000L
alpha <- 0.05
```

# functions
## discretizing mechanisms
### median split
```{r median_split, class.source = "fold-show"}
median_split <- function(data){
  med <- quantile(data$x, 0.5)

  data_med_split <- data |>
    mutate(condition = case_when(
      x <= med ~ 0,
      x >  med ~ 1)) |> 
    mutate(condition = as.factor(condition)) |> 
    select(score = y, condition) 
  
  p_med_split <- get_p_ttest(data_med_split) 
  
  return(p_med_split)
}
```

### tertile split
```{r tertile_split, class.source = "fold-show"}
tertile_split <- function(data){
  
  tertile <- quantile(data$x, c(1/3, 2/3))

  data_tertile_split <- data |> 
    mutate(condition = case_when(
      x <= tertile[1] ~ 0,
      x > tertile[1] & x <= tertile[2] ~ 1,
      x > tertile[2] ~ 2)) |> 
    mutate(condition = as.factor(condition)) |>
    select(score = y, condition)
  
  anova_model <- aov(score ~ condition, data = data_tertile_split)
  p_tertile_split <- summary(anova_model)[[1]][["Pr(>F)"]][[1]] #extract p-value from anova
  
  return(p_tertile_split)
}
```


### cut the middle split
removing the middle categor after a tertile split
```{r cut_middle_split, class.source = "fold-show"}
cut_middle_split <- function(data){
  tertile <- quantile(data$x, c(1/3, 2/3))

  data_cut_middle_split <- data |> 
    mutate(condition = case_when( 
        x <= tertile[1] ~ 0,
        x > tertile[2] ~ 1)) |> 
    mutate(condition = as.factor(condition)) |> 
    select(score = y, condition) |> 
    drop_na() # drops middle tertile which was left NA in condition assignment
  
  p_cut_middle_split <- get_p_ttest(data_cut_middle_split)
  
  return(p_cut_middle_split)
}
```

## discretizing function
```{r discretize_hack, class.source = "fold-show"}
discretize <- function(data){
  p_tibble <- tibble(
    discretizing_mechanism = c(
      "continuous predictor, linear regression",
      "median split predictor, t-test",
      "tertile split predictor, ANOVA",
      "cut the middle split predictor, t-test"
    ),
    p_value = c(
      get_p_univar_linreg(data),
      median_split(data),
      tertile_split(data),
      cut_middle_split(data)
    )
  )
  
  return(p_tibble)
}
```

# simulation
```{r sim_par_discretizing, class.source = "fold-show"}
par_grid_discretizing <- expand_grid(
  n = c(30, 50, 100, 300),
  iteration = 1:n_iterations,
  reporting_strategy = c("first significant","smallest", "smallest significant")
)
```

```{r sim_discretizing, class.source = "fold-show"}
sim_discretizing_file <- paste0("results/sim_discretizing_",n_iterations,"iterations.rds")

if(!file.exists(sim_discretizing_file)){
  set.seed(74392)

  sim_discretizing <- par_grid_discretizing |>
    mutate(
      generated_data = pmap(list(n),
        generate_data_univar_linreg
      ),
      analysis_result = pmap(
        list(generated_data),
        discretize
      ),
      reported_result = pmap(
        list(analysis_result, reporting_strategy, alpha),
        reporting_strategy_tibbles
      )
    )

  saveRDS(sim_discretizing, sim_discretizing_file)
} else {
  sim_discretizing <- readRDS(sim_discretizing_file)
}
```

# results

## replication table
```{r summary_discretizing}
sim_summary_discretizing_file <- paste0("results/sim_summary_discretizing_",n_iterations,"iterations.rds")

if(!file.exists(sim_summary_discretizing_file)){
  sim_summary_discretizing <- sim_discretizing |>
    unnest(reported_result) |>
    group_by(n) |>
    summarize(
      empirical_detection_rate  = mean(p_value < alpha),
      empirical_detection_rate_mcse = sqrt(empirical_detection_rate * (1 - empirical_detection_rate) / n()),
      .groups = "drop"
      )

  saveRDS(sim_summary_discretizing, sim_summary_discretizing_file)
} else {
  sim_summary_discretizing <- readRDS(sim_summary_discretizing_file)
}

sim_summary_discretizing
```

```{r flextable_repli_discretizing}
sim_summary_discretizing |>
  mutate(`n` = as.character(n)) |>
  mutate(across(where(is.numeric), \(x) scales::number(x, accuracy = 0.001))) |>
  mutate(edr_string = paste0(empirical_detection_rate,
                             " (", 
                             empirical_detection_rate_mcse,
                             ")")) |>
  select(`n`,
         `empirical FPR (±1 MCSE)` = edr_string) |>
  apa_flextable(no_format_columns = `n`) |>
  align(j = 1, align = "center") |>
  add_footer_lines(values = as_paragraph(as_i("Note. "), "Empirical False Positive Rate (±1 Monte Carlo Standard Error). n = original total sample size."))
```

```{r FPR_95CI_discretizing}
sim_summary_discretizing |> 
  mutate(FPR_95CI_low = empirical_detection_rate - empirical_detection_rate_mcse * 1.96,
         FPR_95CI_upp = empirical_detection_rate + empirical_detection_rate_mcse * 1.96) |> 
  select(-empirical_detection_rate, -empirical_detection_rate_mcse)
```

## replication plot
```{r plot_repli_discretizing}
ggplot(sim_summary_discretizing, 
       aes(x=n,
           y=empirical_detection_rate)) +
  geom_hline(yintercept=alpha, colour="grey") +
  geom_linerange(aes(ymin = empirical_detection_rate - empirical_detection_rate_mcse,
                     ymax = empirical_detection_rate +empirical_detection_rate_mcse)
                , color = "black"
                ) +
  geom_line() +
  geom_point() +
  labs(
    x = "original sample size",
    y = "empirical false-positive-rate") +
  theme(text = element_text(size=20)) +
  scale_y_continuous(breaks = c(0, 0.05, 0.1, 0.2, 0.3, 0.4), limits = c(0,0.45)) +
  scale_x_continuous(breaks = c(30, 50, 100, 300)) +
  theme_classic()
```
*Note.* Replicated results of the empirical false positive rates when reporting based on $p$-value significance one out of four analyses (linear regression with continuous predictor; three analyses following three different predictor discretization mechanisms). Lineranges show $\pm 1$ Monte Carlo standard error (MCSE) of the empirical false-positive rate and are largely smaller than the plotted points; maximum MCSE across all conditions = .002. The grey line indicates the nominal type I error rate of $\alpha = .05$.

## extension 1: isolated discretizing mechanism
### FPR per mechanism
```{r summary_FPR_mechanism_n_discretizing}
sim_summary_mechanism_n_discretizing <- sim_discretizing |> 
  unnest(analysis_result) |> 
  group_by(n, discretizing_mechanism) |>
  summarize(
    empirical_detection_rate  = mean(p_value < alpha),
    empirical_detection_rate_mcse = sqrt(empirical_detection_rate * (1 - empirical_detection_rate) / n()),
    .groups = "drop"
    )

sim_summary_mechanism_n_discretizing
```

```{r FPR_95CI_isolated_analyses_n}
sim_summary_mechanism_n_discretizing |> 
  mutate(FPR_95CI_low = empirical_detection_rate - empirical_detection_rate_mcse * 1.96,
         FPR_95CI_upp = empirical_detection_rate + empirical_detection_rate_mcse * 1.96) |> 
  select(-empirical_detection_rate, -empirical_detection_rate_mcse)
```

```{r FPR_mechanism_n_flextable_discretizing}
sim_summary_mechanism_n_discretizing |>
  mutate(`n` = as.character(n),
         discretizing_mechanism = as.character(discretizing_mechanism)) |>
  mutate(across(where(is.numeric), \(x) scales::number(x, accuracy = 0.001))) |>
  mutate(edr_string = paste0(empirical_detection_rate,
                             " (", 
                             empirical_detection_rate_mcse,
                             ")")) |>
  select(`n`,
         discretizing_mechanism,
         edr_string) |>
  pivot_wider(names_from = discretizing_mechanism,
              values_from = edr_string) |>
  column_spanner_label("Discretizing mechanism and corresponding analysis", -1) |>
  apa_flextable(no_format_columns = `n`) |>
  align(j = 1, align = "center") |>
  add_footer_lines(values = as_paragraph(as_i("Note. "), "Empirical False Positive Rate (±1 Monte Carlo Standard Error)."))
```

```{r FPR_mechanism_discretizing}
sim_summary_mechanism_discretizing_file <- paste0("results/sim_summary_mechanism_discretizing_",n_iterations,"iterations.rds")

if(!file.exists(sim_summary_mechanism_discretizing_file)){
  sim_summary_mechanism_discretizing <- sim_discretizing |>
    unnest(analysis_result) |>
    group_by(discretizing_mechanism) |>
    summarize(
      empirical_detection_rate  = mean(p_value < alpha),
      empirical_detection_rate_mcse = sqrt(empirical_detection_rate * (1 - empirical_detection_rate) / n()),
      .groups = "drop"
      )

  saveRDS(sim_summary_mechanism_discretizing, sim_summary_mechanism_discretizing_file)
} else {
  sim_summary_mechanism_discretizing <- readRDS(sim_summary_mechanism_discretizing_file)
}

sim_summary_mechanism_discretizing
```

```{r FPR_mechanism_flextable_discretizing}
sim_summary_mechanism_discretizing |>
  mutate(discretizing_mechanism = as.character(discretizing_mechanism)) |>
  mutate(across(where(is.numeric), \(x) scales::number(x, accuracy = 0.001))) |>
  mutate(edr_string = paste0(empirical_detection_rate,
                             " (",
                             empirical_detection_rate_mcse,
                             ")")) |>
  select(`Discretizing mechanism and corresponding analysis` = discretizing_mechanism,
         `empirical FPR (±1 MCSE)` = edr_string) |>
  apa_flextable(no_format_columns = `Discretizing mechanism and corresponding analysis`) |>
  align(j = 1, align = "left") |>
  add_footer_lines(values = as_paragraph(as_i("Note. "), "Empirical False Positive Rate (±1 Monte Carlo Standard Error)."))
```

```{r FPR_95CI_isolated_analyses_}
sim_summary_mechanism_discretizing |> 
  mutate(FPR_95CI_low = empirical_detection_rate - empirical_detection_rate_mcse * 1.96,
         FPR_95CI_upp = empirical_detection_rate + empirical_detection_rate_mcse * 1.96) |> 
  select(-empirical_detection_rate, -empirical_detection_rate_mcse)
```

### count mechanisms within significant FP results
```{r}
discretizing_mechanism_percentage_FPresults <- sim_discretizing |>
  unnest(analysis_result) |>
  filter(p_value < alpha) |> #out of the FP results
  count(discretizing_mechanism) |> # which predictor mechanism was used
  mutate(percentage = n / sum(n))

discretizing_mechanism_percentage_FPresults
```


## extension 2: reporting strategy
### mechanism & reporting
count mechanism within reported results. Depends on reporting strategy and therefore for first significant also on order of application.
```{r}
discretizing_mechanism_percentage_reported_results <- sim_discretizing |>
  unnest(reported_result) |>
  filter(p_value < alpha) |> #out of the FP results
  count(discretizing_mechanism) |> # which predictor mechanism was used
  mutate(percentage = n / sum(n))

discretizing_mechanism_percentage_reported_results
```

```{r}
mechanism_count_reported_results <- sim_discretizing |>
  unnest(reported_result) |>
  filter(p_value < alpha) |>
  count(reporting_strategy, discretizing_mechanism) |>
  group_by(reporting_strategy) |>
  mutate(percentage_within_reporting_strategy = n / sum(n)) |>
  ungroup()

mechanism_count_reported_results
```

### FPR & reporting
```{r summary_FPR_reporting}
sim_discretizing_summary_reporting <- sim_discretizing |> 
  unnest(reported_result) |> 
  group_by(reporting_strategy) |>
  summarize(
    empirical_detection_rate  = mean(p_value < alpha),
    empirical_detection_rate_mcse = sqrt(empirical_detection_rate * (1 - empirical_detection_rate) / n()),
    .groups = "drop"
    )

sim_discretizing_summary_reporting
```

```{r}
sessionInfo()
writeLines(capture.output(sessionInfo()), paste0("results/discretizing_sessionInfo_", n_iterations, "iterations.txt"))
```



