---
title: "discretizing"
author: "Carolin Kirsch"
date: "2026-01-22"
output: html_document
---

```{r}
library(dplyr)
library(tidyverse)
library(tidyr)
library(ggplot2)
library(ggtext)
library(kableExtra)
```

```{r}
source("functions.R")
```

```{r}
n_iterations <- 10000
alpha <- 0.05
```

Split continuous variables into multiple categories using arbitrary cutoff values to achieve statistical significance

Only sample size varied

three mechanisms to discretize a continuous variable: we used a median split, resulting in two groups, a tertile split, resulting in three groups, and a cut-the-middle split, where the middle category is removed after a tertile split to emulate a comparison of extreme groups.

univariate linear regriossion (for the
continuous predictor),
two t-tests (for the median split and the cut the middle strategy),
ANOVA (for the tertile split)

# functions
## data generating
source generate_data_univar_linreg()

## technique
@stefan2023big didn't specify how they calculated the quantiles / if they used a function to split the data. So I assumed they used default settings for stats::quantile().

### median split
```{r}
median_split <- function(data){
  med <- median(data$x)

  data_med_split <- data |> 
  
    mutate(x_med_split = case_when(
      x <= med ~ 0,
      x >  med ~ 1)) |> 
    
    select(y, x_med_split) |> 
    rename(score = y,
           condition = x_med_split)
  
  p_med_split <- get_p_ttest(data_med_split) 
  
  return(p_med_split)
}
```

### tertile split
```{r}
tertile_split <- function(data){
  
  tertile <- quantile(data$x, c(1/3, 2/3))

  data_tertile_split <- data %>%
    mutate(condition = case_when(
      x <= tertile[1] ~ 0,
      x > tertile[1] & x <= tertile[2] ~ 1,
      x > tertile[2] ~ 2)) %>%
    mutate(condition = as.factor(condition)) |>
    select(score = y, condition)
  
  anova_model <- aov(score ~ condition, data = data_tertile_split)
  p_tertile_split <- summary(anova_model)[[1]][["Pr(>F)"]][[1]] #extract pvalue from anova
  
  return(p_tertile_split)
}
```


### cut the middle split
where the middle category is removed after a tertile split to emulate a comparison of extreme groups.
```{r}
cut_middle_split <- function(data){
  tertile <- quantile(data$x, c(1/3, 2/3))

  data_cut_middle_split <- data %>%
    mutate(condition = case_when(
        x <= tertile[1] ~ 0,
        x > tertile[2] ~ 1)) %>%
    select(score = y, condition) |> 
    drop_na()
  
  p_cut_middle_split <- get_p_ttest(data_cut_middle_split)
  
  return(p_cut_middle_split)
}
```

### continous
source get_p_univar_linreg()

## discretizing function

```{r}
discretize <- function(data){
  p_tibble <- tibble(
    p_value_type = c(
      "p_lin_reg",
      "p_med_split",
      "p_tertile_split",
      "p_cut_middle_split"
    ),
    p_value = c(
      get_p_univar_linreg(data),
      median_split(data),
      tertile_split(data),
      cut_middle_split(data)
    )
  )
  
  return(p_tibble)
}
```

## reporting
source reporting_strategy_tibbles()

# simulation
```{r}
experiment_parameters_grid <- expand_grid(
  n = c(30, 50, 100, 300),
  iteration = 1:n_iterations,
  reporting_strategy = c("first significant","smallest", "smallest significant")
)
```

```{r}
set.seed(12345)

simulation <- experiment_parameters_grid |>
  mutate(
    generated_data = pmap(list(n),
      generate_data_univar_linreg
    ),
    analysis_result = pmap(
      list(generated_data),
      discretize
    ),
    reported_result = pmap(
      list(analysis_result, reporting_strategy),
      reporting_strategy_tibbles
    )    
  )
```

# results
## table
```{r}
# summarise simulation results over the iterations ----
simulation_summary <- simulation |>
  unnest(reported_result) |> 
  group_by(n) |>
  summarize(FPR = janitor::round_half_up(mean(p_value < .05), 2))

simulation_summary
```

## plot
```{r}
ggplot(simulation_summary, 
       aes(x=n,
           y=FPR)) +
  geom_line() +
  geom_point() +
  labs(
    x = "n_per_group",
    y = "False-Positive-Rate") +
  ylim(0,0.5) +
  scale_x_continuous(breaks = c(30, 50, 100, 300)) +
  geom_hline(yintercept=alpha, colour="grey") +
  theme_classic()
```
