---
title: "discretizing"
author: "Carolin Kirsch"
date: "2026-01-22"
output:
  html_document: default
  word_document: default
  pdf_document:
    latex_engine: xelatex
---

```{r setup}
library(dplyr)
library(tidyverse)
library(tidyr)
library(ggplot2)
library(ggtext)
library(flextable)
library(apa7)
library(ftExtra)

set_flextable_defaults(theme_fun = theme_apa,
                        font.family = "Times New Roman")
```

```{r source}
source("functions.R")
```

```{r general_settings}
n_iterations <- 10000L
alpha <- 0.05
```

Split continuous variables into multiple categories using arbitrary cutoff values to achieve statistical significance

Only sample size varied

three mechanisms to discretize a continuous variable: we used a median split, resulting in two groups, a tertile split, resulting in three groups, and a cut-the-middle split, where the middle category is removed after a tertile split to emulate a comparison of extreme groups.

univariate linear regriossion (for the continuous predictor),
two t-tests (for the median split and the cut the middle strategy),
ANOVA (for the tertile split)

# functions
## discretizing mechanisms
@stefan2023big didn't specify how they calculated the quantiles / if they used a function to split the data. So I assumed they used default settings for stats::quantile().

### median split
```{r median_split}
median_split <- function(data){
  med <- quantile(data$x, 0.5)

  data_med_split <- data |>
    mutate(condition = case_when(
      x <= med ~ 0,
      x >  med ~ 1)) |> 
    mutate(condition = as.factor(condition)) |> 
    select(score = y, condition) 
  
  p_med_split <- get_p_ttest(data_med_split) 
  
  return(p_med_split)
}
```

### tertile split
```{r tertile_split}
tertile_split <- function(data){
  
  tertile <- quantile(data$x, c(1/3, 2/3))

  data_tertile_split <- data |> 
    mutate(condition = case_when(
      x <= tertile[1] ~ 0,
      x > tertile[1] & x <= tertile[2] ~ 1,
      x > tertile[2] ~ 2)) |> 
    mutate(condition = as.factor(condition)) |>
    select(score = y, condition)
  
  anova_model <- aov(score ~ condition, data = data_tertile_split)
  p_tertile_split <- summary(anova_model)[[1]][["Pr(>F)"]][[1]] #extract p-value from anova
  
  return(p_tertile_split)
}
```


### cut the middle split
where the middle category is removed after a tertile split to emulate a comparison of extreme groups.
```{r cut_middle_split}
cut_middle_split <- function(data){
  tertile <- quantile(data$x, c(1/3, 2/3))

  data_cut_middle_split <- data |> 
    mutate(condition = case_when( 
        x <= tertile[1] ~ 0,
        x > tertile[2] ~ 1)) |> 
    mutate(condition = as.factor(condition)) |> 
    select(score = y, condition) |> 
    drop_na() # drops middle tertile which was left NA in condition assignment
  
  p_cut_middle_split <- get_p_ttest(data_cut_middle_split)
  
  return(p_cut_middle_split)
}
```

## discretizing function
```{r discretize_hack}
discretize <- function(data){
  p_tibble <- tibble(
    discretizing_mechanism = c(
      "continuous predictor, linear regression",
      "median split predictor, t-test",
      "tertile split predictor, ANOVA",
      "cut the middle split predictor, t-test"
    ),
    p_value = c(
      get_p_univar_linreg(data),
      median_split(data),
      tertile_split(data),
      cut_middle_split(data)
    )
  )
  
  return(p_tibble)
}
```

# simulation
```{r sim_par_discretizing}
par_grid_discretizing <- expand_grid(
  n = c(30, 50, 100, 300),
  iteration = 1:n_iterations,
  reporting_strategy = c("first significant","smallest", "smallest significant")
)
```

```{r sim_discretizing}
sim_discretizing_file <- paste0("results/sim_discretizing_",n_iterations,"iterations.rds")

if(!file.exists(sim_discretizing_file)){
  set.seed(74392)

  sim_discretizing <- par_grid_discretizing |>
    mutate(
      generated_data = pmap(list(n),
        generate_data_univar_linreg
      ),
      analysis_result = pmap(
        list(generated_data),
        discretize
      ),
      reported_result = pmap(
        list(analysis_result, reporting_strategy, alpha),
        reporting_strategy_tibbles
      )
    )

  saveRDS(sim_discretizing, sim_discretizing_file)
} else {
  sim_discretizing <- readRDS(sim_discretizing_file)
}
```

# results

## replication table
```{r summary_discretizing}
sim_summary_discretizing_file <- paste0("results/sim_summary_discretizing_",n_iterations,"iterations.rds")

if(!file.exists(sim_summary_discretizing_file)){
  sim_summary_discretizing <- sim_discretizing |>
    unnest(reported_result) |>
    group_by(n) |>
    summarize(
      empirical_detection_rate  = mean(p_value < alpha),
      empirical_detection_rate_mcse = sqrt(empirical_detection_rate * (1 - empirical_detection_rate) / n()),
      .groups = "drop"
      )

  saveRDS(sim_summary_discretizing, sim_summary_discretizing_file)
} else {
  sim_summary_discretizing <- readRDS(sim_summary_discretizing_file)
}

sim_summary_discretizing
```

TO DO: row names, note, caption, etc.
```{r flextable_repli_discretizing}
sim_summary_discretizing |>
  mutate(`n` = as.character(n)) |>
  mutate(across(where(is.numeric), \(x) scales::number(x, accuracy = 0.001))) |>
  mutate(edr_string = paste0(empirical_detection_rate,
                             " (", 
                             empirical_detection_rate_mcse,
                             ")")) |>
  select(`n`,
         `empirical FPR (±1 MCSE)` = edr_string) |>
  #pivot_wider(names_from = p_max,
              #values_from = edr_string) |>
  apa_flextable(no_format_columns = `n`) |>
  align(j = 1, align = "center") |>
  add_footer_lines(values = as_paragraph(as_i("Note. "), "Values in parentheses are ±1 Monte Carlo Standard Error."))
```
<!-- caption for quarto integration: "Empirical False Positive Rate (±1 MCSE) by Baseline Sample Size" -- will be set via #| tbl-cap in the .qmd chunk -->

```{r FPR_95CI_discretizing}
sim_summary_discretizing |> 
  mutate(FPR_95CI_low = empirical_detection_rate - empirical_detection_rate_mcse * 1.96,
         FPR_95CI_upp = empirical_detection_rate + empirical_detection_rate_mcse * 1.96) |> 
  select(-empirical_detection_rate, -empirical_detection_rate_mcse)
```


## replication plot
```{r plot_repli_discretizing}
ggplot(sim_summary_discretizing, 
       aes(x=n,
           y=empirical_detection_rate)) +
  geom_hline(yintercept=alpha, colour="grey") +
  geom_linerange(aes(ymin = empirical_detection_rate - empirical_detection_rate_mcse,
                     ymax = empirical_detection_rate +empirical_detection_rate_mcse)
                , color = "black"
                ) +
  geom_line() +
  geom_point() +
  labs(
    x = "original sample size",
    y = "empirical false-positive-rate") +
  theme(text = element_text(size=20)) +
  scale_y_continuous(breaks = c(0, 0.05, 0.1, 0.2, 0.3, 0.4), limits = c(0,0.45)) +
  scale_x_continuous(breaks = c(30, 50, 100, 300)) +
  theme_classic()
```
\textit{Note.} Replicated results of the empirical false positive rates when reporting based on $p$-value significance one out of four analyses (linear regression with continuous predictor; three analyses following three different predictor discretization mechanisms). <!--TO DO chunky description that is maybe to close to OG as well -->. Lineranges would show $\pm 1$ Monte Carlo standard error (MCSE) of the emprical false-positive- rate; maximum MCSE across all conditions = .002. The grey line indicates the nominal type I error rate of $\alpha = .05$.

## extension 1: isolated discretizing mechanism
### FPR per mechanism
```{r summary_FPR_mechanism_n_discretizing}
sim_summary_mechanism_n_discretizing <- sim_discretizing |> 
  unnest(analysis_result) |> 
  group_by(n, discretizing_mechanism) |>
  summarize(
    empirical_detection_rate  = mean(p_value < alpha),
    empirical_detection_rate_mcse = sqrt(empirical_detection_rate * (1 - empirical_detection_rate) / n()),
    .groups = "drop"
    )

sim_summary_mechanism_n_discretizing
```

```{r FPR_95CI_isolated_analyses_n}
sim_summary_mechanism_n_discretizing |> 
  mutate(FPR_95CI_low = empirical_detection_rate - empirical_detection_rate_mcse * 1.96,
         FPR_95CI_upp = empirical_detection_rate + empirical_detection_rate_mcse * 1.96) |> 
  select(-empirical_detection_rate, -empirical_detection_rate_mcse)
```

TO DO: row names, note, caption, etc.
```{r FPR_mechanism_n_flextable_discretizing}
sim_summary_mechanism_n_discretizing |>
  mutate(`n` = as.character(n),
         discretizing_mechanism = as.character(discretizing_mechanism)) |>
  mutate(across(where(is.numeric), \(x) scales::number(x, accuracy = 0.001))) |>
  mutate(edr_string = paste0(empirical_detection_rate,
                             " (", 
                             empirical_detection_rate_mcse,
                             ")")) |>
  select(`n`,
         discretizing_mechanism,
         edr_string) |>
  pivot_wider(names_from = discretizing_mechanism,
              values_from = edr_string) |>
  column_spanner_label("Discretizing mechanism and corresponding analysis", -1) |>
  apa_flextable(no_format_columns = `n`) |>
  align(j = 1, align = "center") |>
  add_footer_lines(values = as_paragraph(as_i("Note. "), "Values in parentheses are ±1 Monte Carlo Standard Error."))
```
<!-- caption for quarto integration: "Empirical False Positive Rate (±1 MCSE) by Original Sample Size and Discretizing Mechanism" -- will be set via #| tbl-cap in the .qmd chunk -->


```{r FPR_mechanism_discretizing}
sim_summary_mechanism_discretizing_file <- paste0("results/sim_summary_mechanism_discretizing_",n_iterations,"iterations.rds")

if(!file.exists(sim_summary_mechanism_discretizing_file)){
  sim_summary_mechanism_discretizing <- sim_discretizing |>
    unnest(analysis_result) |>
    group_by(discretizing_mechanism) |>
    summarize(
      empirical_detection_rate  = mean(p_value < alpha),
      empirical_detection_rate_mcse = sqrt(empirical_detection_rate * (1 - empirical_detection_rate) / n()),
      .groups = "drop"
      )

  saveRDS(sim_summary_mechanism_discretizing, sim_summary_mechanism_discretizing_file)
} else {
  sim_summary_mechanism_discretizing <- readRDS(sim_summary_mechanism_discretizing_file)
}

sim_summary_mechanism_discretizing
```

TO DO: row names, note, caption, etc.
```{r FPR_mechanism_flextable_discretizing}
sim_summary_mechanism_discretizing |>
  mutate(discretizing_mechanism = as.character(discretizing_mechanism)) |>
  mutate(across(where(is.numeric), \(x) scales::number(x, accuracy = 0.001))) |>
  mutate(edr_string = paste0(empirical_detection_rate,
                             " (",
                             empirical_detection_rate_mcse,
                             ")")) |>
  select(`Discretizing mechanism and corresponding analysis` = discretizing_mechanism,
         `empirical FPR (±1 MCSE)` = edr_string) |>
  apa_flextable(no_format_columns = `Discretizing mechanism and corresponding analysis`) |>
  align(j = 1, align = "left") |>
  add_footer_lines(values = as_paragraph(as_i("Note. "), "Values in parentheses are ±1 Monte Carlo Standard Error."))
```
<!-- caption for quarto integration: "Empirical False Positive Rate (±1 MCSE) by Discretizing Mechanism" -- will be set via #| tbl-cap in the .qmd chunk -->


```{r FPR_95CI_isolated_analyses_}
sim_summary_mechanism_discretizing |> 
  mutate(FPR_95CI_low = empirical_detection_rate - empirical_detection_rate_mcse * 1.96,
         FPR_95CI_upp = empirical_detection_rate + empirical_detection_rate_mcse * 1.96) |> 
  select(-empirical_detection_rate, -empirical_detection_rate_mcse)
```

### count mechanisms within significant FP results
```{r}
discretizing_mechanism_percentage_FPresults <- sim_discretizing |>
  unnest(analysis_result) |>
  filter(p_value < alpha) |> #out of the FP results
  count(discretizing_mechanism) |> # which predictor mechanism was used
  mutate(percentage = n / sum(n))

discretizing_mechanism_percentage_FPresults
```


## extension 2: reporting strategy

### mechanism & reporting
has to do with a) order of implementation for first significant and b) with p-value distribution for the other two (I think at leas. So currently I assume that median split does lead to smaller p-values. Possibly has to do with power? at least it has the most power out of the discretizing mechanisms. For the comparisons with baseline I am not sure what the mechanism is)

TO DO: 1) decide whether to keep or not 2) maybe rename "n"?
```{r}
discretizing_mechanism_percentage_reported_results <- sim_discretizing |>
  unnest(reported_result) |>
  filter(p_value < alpha) |> #out of the FP results
  count(discretizing_mechanism) |> # which predictor mechanism was used
  mutate(percentage = n / sum(n))

discretizing_mechanism_percentage_reported_results
```

```{r}
mechanism_count_reported_results <- sim_discretizing |>
  unnest(reported_result) |>
  filter(p_value < alpha) |>
  count(reporting_strategy, discretizing_mechanism) |>
  group_by(reporting_strategy) |>
  mutate(percentage_within_reporting_strategy = n / sum(n)) |>
  ungroup()

mechanism_count_reported_results
```

TO DO: row names, note, caption, etc.
TO DO: would need different order to be readable
```{r mechanism_count_FPR_flextable_discretizing}
mechanism_count_reported_results |>
  mutate(`reporting strategy` = as.character(reporting_strategy),
         discretizing_mechanism = as.factor(discretizing_mechanism)) |>
  mutate(across(where(is.numeric), \(x) scales::number(x, accuracy = 0.001))) |>
  select(`reporting strategy`,
         discretizing_mechanism,
         percentage_within_reporting_strategy) |>
  pivot_wider(names_from = discretizing_mechanism,
              values_from = percentage_within_reporting_strategy) |>
  column_spanner_label("Discretizing mechanism and corresponding analysis", -1) |>
  apa_flextable(no_format_columns = `reporting strategy`) |>
  align(j = 1, align = "left") |>
  add_footer_lines(values = as_paragraph(as_i("Note.")))
```
<!-- caption for quarto integration: "Percentages of Discretizing Mechanisms within Reported False-Positive Analysis Results" -- will be set via #| tbl-cap in the .qmd chunk -->


### FPR & reporting
```{r summary_FPR_reporting}
sim_discretizing_summary_reporting <- sim_discretizing |> 
  unnest(reported_result) |> 
  group_by(reporting_strategy) |>
  summarize(
    empirical_detection_rate  = mean(p_value < alpha),
    empirical_detection_rate_mcse = sqrt(empirical_detection_rate * (1 - empirical_detection_rate) / n()),
    .groups = "drop"
    )

sim_discretizing_summary_reporting
```

```{r}
sessionInfo()
```

```{r}
writeLines(capture.output(sessionInfo()), paste0("results/discretizing_sessionInfo_", n_iterations, "iterations.txt"))
```


