---
title: "incorrect rounding"
author: "Carolin Kirsch"
date: "2025-12-27"
output: html_document
---

```{r}
library(dplyr)
library(tidyverse)
library(tidyr)
library(ggplot2)
library(ggtext)
library(kableExtra)
```

```{r}
source("functions.R")
```

```{r}
n_iterations <- 30000 # 30'000 since no reporting strategy that is varied
alpha <- 0.05 #for exploring alterantive sig. levels the thresholds for rounding down (p_max) would need to be adapted as well
```

Round p-value down to the significance level. The maximum value $p_{max}$ that is rounded down to $\alpha = 0.05$ is varied with $p_{max} \in \{0.051, 0.6\}$.

# functions
## generate data

OPEN: decide if I want to change structure to score, condition. Just for consistency reasons.
```{r}
generate_data <- function(n,
                          mu = 0,
                          sd = 1){
  x <- rnorm(n, mu, sd)
  y <- rnorm(n, mu, sd)

  data <- tibble(x=x, y=y)
  
  return(data)
}
```

## analyse data
OPEN: update to use data_generating_ttest() ?

```{r}
analyse_data <- function(data){
  
  result <- t.test(data$x,data$y, 
                   alternative = "two.sided", 
                   paired = FALSE, 
                   var.equal = TRUE)
  
  p <- result$p.value
  
  return(p)
}
```

## incorrect rounding
```{r}
incorrect_rounding <- function(p, p_max){
  ifelse(alpha < p & p <= p_max, 
         p_hacked <- alpha,
         p_hacked <- p) 
  
  return(p_hacked)
}
```

## reporting strategies
doesn't matter here since there are no multiple tests run.

# simulation
```{r}
experiment_parameters_grid <- expand_grid(
  n = c(30, 50, 100, 300),
  iteration = 1:n_iterations,
  p_max = c(0.051, 0.06),
  reporting_strategy = c("not applicable"))
```

```{r}
set.seed(12345) 

simulation <- 

  experiment_parameters_grid |>
  
  mutate(generated_data = pmap(list(n),
                               generate_data)) |>
  
  mutate(analysis_results = pmap(list(generated_data),
                                 analyse_data)) |> 
  
  mutate(p_hacked = pmap(list(analysis_results, p_max),
                  incorrect_rounding))
  
```

# results
## table 
```{r}
simulation_summary <- simulation |>
  mutate(n = as.factor(n),
         p_max = as.factor(p_max)) |> 
  group_by(n, p_max) |>
  dplyr::summarize(false_positive_rate = mean(p_hacked <= .05), .groups = "drop")

simulation_summary
```

## plot
OPEN: how do I want to plot this? The original didn't include a plot on it.
```{r}
ggplot(simulation_summary,
       aes(x = p_max,
           y = false_positive_rate,
           group = n,
           color = n)) +
  geom_line() +
  geom_point() +
  labs(
    x = "p_max",
    y = "False-Positive-Rate",
    color = "n"
  ) +
  theme_minimal()
```

```{r}
ggplot(simulation_summary,
       aes(x = n,
           y = false_positive_rate,
           group = p_max,
           color = p_max)) +
  geom_line() +
  geom_point() +
  labs(
    x = "sample size",
    y = "False-Positive-Rate",
    color = "p_max"
  ) +
  theme_minimal()
```
