---
title: "Incorrect Rounding"
author: "Carolin Kirsch"
output:
  html_document:
    code_folding: hide
---

```{r setup, message = FALSE, warning = FALSE, class.source = "fold-show"}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(dplyr)
library(tidyverse)
library(tidyr)
library(ggplot2)
library(flextable)
library(apa7)
library(ftExtra)

set_flextable_defaults(theme_fun = theme_apa,
                        font.family = "Times New Roman")
```

```{r source, class.source = "fold-show"}
source("functions.R")
```

```{r general_settings, class.source = "fold-show"}
n_iterations <-  30000L # 30'000 since no reporting strategy that is varied
alpha <- 0.05 #for exploring alternative sig. levels the thresholds for rounding down (p_max) would need to be adapted as well
```

# functions
## incorrect rounding
```{r incorrect_rounding, class.source = "fold-show"}
incorrect_rounding <- function(p, p_max){
  ifelse(alpha < p & p <= p_max, 
         p_hacked <- alpha,
         p_hacked <- p) 
  
  return(p_hacked)
}
```

# simulation
```{r sim_par_rounding, class.source = "fold-show"}
par_grid_rounding <- expand_grid(
  n = c(30, 50, 100, 300),
  iteration = 1:n_iterations,
  p_max = c(0.051, 0.06),
  reporting_strategy = c("not applicable"))
```

```{r sim_rounding, class.source = "fold-show"}
rounding_cache_file <- paste0("results/sim_rounding_", n_iterations, "iterations.rds")

if (file.exists(rounding_cache_file)) {
  sim_rounding <- readRDS(rounding_cache_file)
} else {
  set.seed(12345)

  sim_rounding <- par_grid_rounding |>
    mutate(
      generated_data = pmap(list(n),
                            generate_data_ttest),

      analysis_results = pmap(list(generated_data),
                              get_p_ttest),

      p_hacked = pmap(list(analysis_results, p_max),
                      incorrect_rounding)
      )

  saveRDS(sim_rounding, rounding_cache_file)
}
```

# results
## table 
```{r rounding_summary}
sim_summary_rounding_file <- paste0("results/sim_summary_rounding_",n_iterations,"iterations.rds")

if(!file.exists(sim_summary_rounding_file)){
  sim_summary_rounding <- sim_rounding |>
    group_by(n, p_max) |>
    summarize(
      empirical_detection_rate  = mean(p_hacked <= 0.05), #inclusive rejection criterion!
      empirical_detection_rate_mcse = sqrt(empirical_detection_rate * (1 - empirical_detection_rate) / n()),
      .groups = "drop"
      )

  saveRDS(sim_summary_rounding, sim_summary_rounding_file)
} else {
  sim_summary_rounding <- readRDS(sim_summary_rounding_file)
}

sim_summary_rounding
```

```{r rounding_flextable_table}
sim_summary_rounding |>
  mutate(`n` = as.character(n),
         p_max = as.character(p_max)) |>
  mutate(across(where(is.numeric), \(x) scales::number(x, accuracy = 0.001))) |>
  mutate(edr_string = paste0(empirical_detection_rate,
                             " (", 
                             empirical_detection_rate_mcse,
                             ")")) |>
  select(`n`,
         p_max,
         edr_string) |>
  pivot_wider(names_from = p_max,
              values_from = edr_string) |>
  column_spanner_label("Rounding threshold", -1) |>
  apa_flextable(no_format_columns = `n`) |>
  align(j = 1, align = "center") |>
  add_footer_lines(values = as_paragraph(as_i("Note. "), "Empirical False Positive Rate (±1 Monte Carlo Standard Error). n = sample size per group."))
```


## plot
```{r plot_repli}
ggplot(sim_summary_rounding,
       aes(x = p_max,
           y = empirical_detection_rate,
           color = as.factor(n))) +
  geom_hline(yintercept = 0.05, col = "grey") +
  geom_linerange(aes(ymin = empirical_detection_rate - empirical_detection_rate_mcse,
                     ymax = empirical_detection_rate +empirical_detection_rate_mcse)
                , color = "black") +
  geom_line() +
  geom_point() +
  scale_y_continuous(breaks = c(0, 0.05, 0.1, 0.2, 0.3, 0.4), limits = c(0,0.45)) +
  scale_x_continuous(breaks = c(0.051, 0.06)) +
  labs(
    x = "p_max",
    y = "empirical false-positive-rate",
    color = "n per group") +
  theme_classic()
```

*Note.* Lineranges show ±1 Monte Carlo standard error (MCSE) of the empirical false-positive rate and are largely smaller than the plotted points; maximum MCSE across all conditions = .001.The grey line indicates the nominal $\alpha$-level of .05. $n$ = sample size per group.

```{r plot_samplesize_x}
ggplot(sim_summary_rounding,
       aes(x = n,
           y = empirical_detection_rate,
           color = as.factor(p_max))) +
  geom_hline(yintercept = 0.05, col = "grey") +
  geom_linerange(aes(ymin = empirical_detection_rate - empirical_detection_rate_mcse,
                     ymax = empirical_detection_rate +empirical_detection_rate_mcse)
                , color = "black") +
  geom_line() +
  geom_point() +
  labs(
    x = "sample size per group",
    y = "empirical false-positive-rate",
    color = "p_max"
  ) +
  theme(text = element_text(size=20)) +
  scale_y_continuous(breaks = c(0, 0.05, 0.1, 0.2, 0.3, 0.4), limits = c(0,0.45)) +
  scale_x_continuous(breaks = c(30, 50, 100, 300)) +
  theme_classic()
```

*Note.* Lineranges show ±1 Monte Carlo standard error (MCSE) of the empirical false-positive rate and are largely smaller than the plotted points; maximum MCSE across all conditions = .001.

```{r}
sessionInfo()

writeLines(capture.output(sessionInfo()), "results/rounding_sessionInfo.txt")
```

