---
title: "optional stopping"
author: "Carolin Kirsch"
date: "2025-12-28"
output: html_document
---

```{r}
library(tidyverse)
library(ggtext)
library(kableExtra)
library(furrr)
plan(multisession, workers = parallel::detectCores() - 1)
```

```{r}
source("functions.R")
```

```{r}
n_iterations <- 5000 # 30'000 since no reporting strategy that is varied
alpha <- 0.05 # hmmm i kinda want to specify the default in the function? should i still do this here then? not really consistent now
```

Conduct intermediate significance tests during data collection, stop data collection when $p < \alpha$. Two versions:

(1) Minimum sample size fixed: N_min = 5; Maximum sample size varied: N_max ∈ {30,50,100,300}; Step size: k ∈ {1,5,10,50}
(2) Minimum sample size varied: N_min ∈ {5, 30, 50, 100}; Maximum sample size fixed: N_max = 300; Step size: k ∈ {1,5,10,50}

If step size > N_max - N_min:
simulate two peeks (at the minimum and the maximum sample size)

# functions
## data generating
generate_data_ttest() sourced from functions file. 

## analysis -> p value extraction
get_p_ttest() sourced from functions file.

## p-hacking

```{r}
optional_stopping <- function(data, n_min, n_max, k, alpha = 0.05){
  
  # define sample sizes at which a peek takes place
  #in case of k > n_max-n_min -> two peeks at (n_min, n_max)
  n_peek <- seq(n_min, n_max,by = k) 
  if(length(n_peek) == 1) { 
    n_peek <- c(n_min, n_max)
  }
  
  count_peeks <- 0L
  p <- NA_real_
 
  # for loop necessary since I need the break criterion
  # this drastically reduces computational effort 
  for(n in n_peek){
    count_peeks <- count_peeks + 1L
    
    p <- data |> # data is generated with sample size n_max
      group_by(condition) |> # then out of each condition
      slice(1:n) |> # the currently tested sample size gets extracted
      ungroup() |>
      get_p_ttest()
    
    # stopping when significant (-> equivalent to reporting strategy first. sig.)
    if(p < alpha) break 
  }

  res <- tibble(
    #hacked_data = list(data), #saving computational power
    count_peeks = count_peeks,
    n_stop = n_peek[count_peeks],
    p = p)
  
  return(res)
}
```

## reporting strategy
doesn't apply here. "First significant" is already inherently part of the p-hacking strategy.

# simulation
## simulation n_max varied
and n_min fixed
```{r}
par_grid_optstop_nmax <- expand_grid(
  iteration = 1:n_iterations,
  n_min = 5,
  n_max = c(30,50,100,300),
  k = c(1,5,10,50),
  reporting_strategy = "not applicable" #OPEN: maybe leave out completely? especially if I am not analyzing it at all later
)
```

```{r}
set.seed(12345)

sim_optstop_nmax <- par_grid_optstop_nmax |>
  mutate(
    generated_data = pmap(
      list(n_max),
      generate_data_ttest
    ),
    analysis_result = future_pmap(   # parallelising
      .l = list(generated_data, n_min, n_max, k), 
      .f = optional_stopping,
      .progress = TRUE,
      .options = furrr_options(seed = TRUE)
    )
  )
```


```{r eval=FALSE, include=FALSE}
set.seed(12345)
 id=as.integer(Sys.getenv("SLURM_ARRAY_TASK_ID"))
sim_optstop_nmax <- par_grid_optstop_nmax |>
  mutate(
    generated_data = pmap(
      list(n_max),
      generate_data_ttest
    ),
    analysis_result = future_pmap(   # parallelising
      .l = list(generated_data[id,], n_min, n_max, k), 
      .f = optional_stopping,
      .progress = TRUE,
      .options = furrr_options(seed = TRUE)
    )
  )

saveRDS(sim_optstop_nmax, paste0("results/sim_optstop_nmax_",id,".rds"))
```

## simulation n_min varied
n_min varied
```{r sim_par_optstop}
par_grid_optstop_nmin <- expand_grid(
  iteration = 1:n_iterations,
  n_min = c(5, 30, 50, 100),
  n_max = 300,
  k = c(1,5,10,50),
  reporting_strategy = "not applicable"
)
```

```{r sim_optstop}
set.seed(12345)

sim_optstop_nmin <- par_grid_optstop_nmin |>
  mutate(
    generated_data = pmap(
      list(n_max),
      generate_data_ttest
    ),
    analysis_result = future_pmap(   # parallelising
      .l = list(generated_data, n_min, n_max, k),
      .f = optional_stopping,
      .progress = TRUE,
      .options = furrr_options(seed = TRUE)
    )
  )
```

```{r}

saveRDS(sim_optstop_nmax, paste0("results/sim_optstop_nmax_",n_iterations,"iterations.rds"))
saveRDS(sim_optstop_nmin, paste0("results/sim_optstop_nmin_",n_iterations,"iterations.rds"))
```

# results
## table 
```{r}
sim_optstop_nmax_summary <- sim_optstop_nmax |>
  unnest(analysis_result) |> 
  mutate(n_max = as.factor(n_max)#,
         #k = as.factor(k)
         ) |> 
  group_by(n_max, k) |>
  dplyr::summarize(false_positive_rate = mean(p < alpha), .groups = "drop")

sim_optstop_nmax_summary
```

```{r}
sim_optstop_nmin_summary <- sim_optstop_nmin |>
  unnest(analysis_result) |> 
  mutate(n_min = as.factor(n_min)#,
         #k = as.factor(k)
         ) |> 
  group_by(n_min, k) |>
  dplyr::summarize(false_positive_rate = mean(p < alpha), .groups = "drop")

sim_optstop_nmin_summary
```

OPEN: potentially add number of hypothetical peeks? since they are actually what influence the FPR
decide later. when I decide what I want to show additionally
```{r eval=FALSE, include=FALSE}
sim_optstop_nmax |>
  mutate(max_peeks = length(seq(n_min, n_max, k)))
#something something table 
```

## plots
```{r}
ggplot(sim_optstop_nmax_summary,
       aes(x = k,
           y = false_positive_rate,
           group = n_max,
           color = n_max)) +
  geom_line() +
  geom_point() +
  labs(
    x = "step size",
    y = "False-Positive-Rate",
    color = "n_max"
  ) +
  theme_minimal()
```

```{r}
ggplot(sim_optstop_nmin_summary,
       aes(x = k,
           y = false_positive_rate,
           group = n_min,
           color = n_min)) +
  geom_line() +
  geom_point() +
  labs(
    x = "step size",
    y = "False-Positive-Rate",
    color = "n_min"
  ) +
  theme_minimal()
```

