---
title: "Optional Stopping"
author: "Carolin Kirsch"
output:
  html_document:
    code_folding: hide
  word_document: default
  pdf_document:
    latex_engine: xelatex
---

```{r setup, message = FALSE, warning = FALSE, class.source = "fold-show"}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

library(tidyverse)
library(flextable)
library(apa7)
library(ftExtra)

set_flextable_defaults(theme_fun = theme_apa,
                        font.family = "Times New Roman")
```

```{r general_settings, class.source = "fold-show"}
# fixed design constants of the UBELIX run (must match optstop_function_run_sim.R;
# checked against the raw data below whenever the raw data is loaded)
n_iterations <- 30000L
alpha <- 0.05
```

Note: the raw per-iteration outputs (`results_UBELIX/sim_optstop_*_30000iterations.rds`) are large, git-ignored and NOT distributed. They are only loaded when one of the cached summary files below is missing; regenerate them with `optstop_function_run_sim.R` in that case. With all summary files present (as shipped), this document knits without the raw data.

```{r load_results, class.source = "fold-show"}
sim_optstop_nmax_summary_file <- paste0("results_UBELIX/sim_optstop_nmax_summary_",n_iterations,"iterations.rds")
sim_optstop_nmin_summary_file <- paste0("results_UBELIX/sim_optstop_nmin_summary_",n_iterations,"iterations.rds")
optstop_peek_counts_nmax_file <- paste0("results_UBELIX/sim_optstop_nmax_peek_counts_summary_",n_iterations,"iterations.rds")
optstop_peek_counts_nmin_file <- paste0("results_UBELIX/sim_optstop_nmin_peek_counts_summary_",n_iterations,"iterations.rds")
optstop_peeks_at_stop_nmax_file <- paste0("results_UBELIX/sim_optstop_nmax_peeks_at_stop_summary_",n_iterations,"iterations.rds")

raw_needed <- !all(file.exists(c(sim_optstop_nmax_summary_file,
                                 sim_optstop_nmin_summary_file,
                                 optstop_peek_counts_nmax_file,
                                 optstop_peek_counts_nmin_file,
                                 optstop_peeks_at_stop_nmax_file)))

if (raw_needed) {
  sim_optstop_nmax <- readRDS(paste0("results_UBELIX/sim_optstop_nmax_",n_iterations,"iterations.rds"))
  sim_optstop_nmin <- readRDS(paste0("results_UBELIX/sim_optstop_nmin_",n_iterations,"iterations.rds"))

  # check the constants above against the raw data
  stopifnot(max(sim_optstop_nmax$iteration) == n_iterations,
            sim_optstop_nmax$alpha[1] == alpha)

  sim_optstop_nmin <- bind_rows(
    sim_optstop_nmin,
    # add the four conditions that shared between the sets and only estimated once
    sim_optstop_nmax |> filter(n_max == 300)
  )
}
```

# replication
## table nmax
Note: this chunk only recomputes and saves the summary if `results_UBELIX/sim_optstop_nmax_summary_<n_iterations>iterations.rds` does not already exist; otherwise the existing file is loaded.
```{r summary_nmax}
if (!file.exists(sim_optstop_nmax_summary_file)) {
  sim_optstop_nmax_summary <- sim_optstop_nmax |>
    unnest(reported_result) |>
    group_by(n_max, k, count_peeks_hypothetical) |>
    summarize(
      empirical_detection_rate  = mean(p_value < alpha),
      empirical_detection_rate_mcse = sqrt(empirical_detection_rate * (1 - empirical_detection_rate) / n()),
      .groups = "drop"
      )

  saveRDS(sim_optstop_nmax_summary, sim_optstop_nmax_summary_file)
} else {
  sim_optstop_nmax_summary <- readRDS(sim_optstop_nmax_summary_file)
}

sim_optstop_nmax_summary
```

```{r nmax_flextable_table}
sim_optstop_nmax_summary |>
  mutate(`maximum n per condition` = as.character(n_max),
         `step size` = as.character(k)) |>
  mutate(across(where(is.numeric), \(x) scales::number(x, accuracy = 0.001))) |>
  mutate(edr_string = paste0(empirical_detection_rate,
                             " (", 
                             empirical_detection_rate_mcse,
                             ")")) |>
  select(`maximum n per condition`,
         `step size`,
         edr_string) |>
  pivot_wider(names_from = `step size`,
              values_from = edr_string) |>
  column_spanner_label("step size", -1) |>
  apa_flextable(no_format_columns = `maximum n per condition`) |>
  align(j = 1, align = "center") |>
  add_footer_lines(values = as_paragraph(as_i("Note. "), "Empirical False Positive Rate (±1 Monte Carlo Standard Error)."))
```
<!-- caption for quarto integration: "Empirical False Positive Rate (±1 MCSE) by Maximum Sample Size and Step Size" -- will be set via #| tbl-cap in the .qmd chunk -->

## table nmin
Note: this chunk only recomputes and saves the summary if `results_UBELIX/sim_optstop_nmin_summary_<n_iterations>iterations.rds` does not already exist; otherwise the existing file is loaded.
```{r summary_nmin}
if (!file.exists(sim_optstop_nmin_summary_file)) {
  sim_optstop_nmin_summary <- sim_optstop_nmin |>
    unnest(reported_result) |>
    group_by(n_min, k, count_peeks_hypothetical) |>
    summarize(
      empirical_detection_rate  = mean(p_value < alpha),
      empirical_detection_rate_mcse = sqrt(empirical_detection_rate * (1 - empirical_detection_rate) / n()),
      .groups = "drop"
      )

  saveRDS(sim_optstop_nmin_summary, sim_optstop_nmin_summary_file)
} else {
  sim_optstop_nmin_summary <- readRDS(sim_optstop_nmin_summary_file)
}

sim_optstop_nmin_summary
```

```{r nmin_flextable_table}
sim_optstop_nmin_summary |>
  mutate(`initial n per condition` = as.character(n_min),
         `step size` = as.character(k)) |>
  mutate(across(where(is.numeric), \(x) scales::number(x, accuracy = 0.001))) |>
  mutate(edr_string = paste0(empirical_detection_rate,
                             " (", 
                             empirical_detection_rate_mcse,
                             ")")) |>
  select(`initial n per condition`,
         `step size`,
         edr_string) |>
  pivot_wider(names_from = `step size`,
              values_from = edr_string) |>
  column_spanner_label("step size", -1) |>
  apa_flextable(no_format_columns = `initial n per condition`) |>
  align(j = 1, align = "center") |>
  add_footer_lines(values = as_paragraph(as_i("Note. "), "Empirical False Positive Rate (±1 Monte Carlo Standard Error)."))
```
<!-- caption for quarto integration: "Empirical False Positive Rate (±1 MCSE) by Initial Sample Size and Step Size" -- will be set via #| tbl-cap in the .qmd chunk -->

## plot nmax
```{r nmax_plot_repli}
ggplot(sim_optstop_nmax_summary,
       aes(x = k,
           y = empirical_detection_rate,
           color = as.factor(n_max))) +
  geom_hline(yintercept = alpha, col = "grey") +
  geom_linerange(aes(ymin = empirical_detection_rate - empirical_detection_rate_mcse,
                     ymax = empirical_detection_rate + empirical_detection_rate_mcse)
                , color = "black") +
  geom_line() +
  geom_point() +
  geom_text(aes(label = count_peeks_hypothetical),
            nudge_x = 1, vjust = -0.8, size = 3,
            show.legend = FALSE) +
  #scale_y_continuous(breaks = c(0, 0.05, 0.1, 0.2, 0.3, 0.4), limits = c(0,0.45)) +
  scale_x_continuous(breaks = c(1, 5, 10, 50)) +
  labs(
    x = "step size",
    y = "empirical false-positive-rate",
    color = "maximum n per group") +
  theme_classic()
```

*Note.* Lineranges show ±1 Monte Carlo standard error (MCSE) of the empirical false-positive rate and are largely smaller than the plotted points; maximum MCSE across all conditions = .003.

## plot nmin
```{r nmin_plot_repli}
ggplot(sim_optstop_nmin_summary,
       aes(x = k,
           y = empirical_detection_rate,
           color = as.factor(n_min))) +
  geom_hline(yintercept = alpha, col = "grey") +
  geom_linerange(aes(ymin = empirical_detection_rate - empirical_detection_rate_mcse,
                     ymax = empirical_detection_rate + empirical_detection_rate_mcse)
                , color = "black") +
  geom_line() +
  geom_point() +
  geom_text(aes(label = count_peeks_hypothetical),
            nudge_x = 1, vjust = -0.8, size = 3,
            show.legend = FALSE) +
  #scale_y_continuous(breaks = c(0, 0.05, 0.1, 0.2, 0.3, 0.4), limits = c(0,0.45)) +
  scale_x_continuous(breaks = c(1, 5, 10, 50)) +
  labs(
    x = "step size",
    y = "empirical false-positive-rate",
    color = "initial n per group") +
  theme_classic()
```

*Note.* Lineranges show ±1 Monte Carlo standard error (MCSE) of the empirical false-positive rate and are largely smaller than the plotted points; maximum MCSE across all conditions = .003.

# extension
## span of FPR
```{r nmin_summary_extension}
sim_optstop_nmin_summary |> 
  group_by(k) |>
  summarize(span = max(empirical_detection_rate) - min(empirical_detection_rate))
```

```{r nmax_summary_extension}
sim_optstop_nmax_summary |> 
  group_by(k) |>
  summarize(FPR_span = max(empirical_detection_rate) - min(empirical_detection_rate))
```

## hypothetical peeks
### count hypothetical peeks
Note: these chunks only recompute and save the peek-count summaries from the raw data if the cached file does not already exist; otherwise the existing file is loaded.
```{r peek_counts_nmax}
if (!file.exists(optstop_peek_counts_nmax_file)) {
  optstop_peek_counts_nmax <- sim_optstop_nmax |>
    unnest(reported_result) |>
    group_by(k) |>
    count(count_peeks_hypothetical, .drop = FALSE) |>
    mutate(n = n/n_iterations) |>
    rename(count = n)

  saveRDS(optstop_peek_counts_nmax, optstop_peek_counts_nmax_file)
} else {
  optstop_peek_counts_nmax <- readRDS(optstop_peek_counts_nmax_file)
}

optstop_peek_counts_nmax
```


```{r peek_counts_nmin}
if (!file.exists(optstop_peek_counts_nmin_file)) {
  optstop_peek_counts_nmin <- sim_optstop_nmin |>
    unnest(reported_result) |>
    group_by(k) |>
    count(count_peeks_hypothetical, .drop = FALSE) |>
    mutate(n = n/n_iterations) |>
    rename(count = n)

  saveRDS(optstop_peek_counts_nmin, optstop_peek_counts_nmin_file)
} else {
  optstop_peek_counts_nmin <- readRDS(optstop_peek_counts_nmin_file)
}

optstop_peek_counts_nmin
```

### linear x-axis
```{r}
ggplot(sim_optstop_nmax_summary,
       aes(x = count_peeks_hypothetical, y = empirical_detection_rate)) +
  geom_hline(yintercept = alpha, colour = "grey") +
  geom_linerange(aes(ymin = empirical_detection_rate - empirical_detection_rate_mcse,
                     ymax = empirical_detection_rate + empirical_detection_rate_mcse)
                , color = "black") +
  geom_line(aes(colour = as.factor(k), group = as.factor(k))) +
  geom_point(aes(colour = as.factor(k), shape = as.factor(n_max)), size = 2) +
  labs(x = "hypothetical number of peeks",
       y = "empirical false positive rate",
       colour = "step size",
       shape = "upper bound for n") +
  theme_classic()
```

*Note.* Lineranges show ±1 Monte Carlo standard error (MCSE) of the empirical false-positive rate and are largely smaller than the plotted points; maximum MCSE across all conditions = .003.

```{r}
ggplot(sim_optstop_nmax_summary,
       aes(x = count_peeks_hypothetical, y = empirical_detection_rate)) +
    geom_hline(yintercept = alpha, colour = "grey") +
    geom_linerange(aes(ymin = empirical_detection_rate - empirical_detection_rate_mcse,
                       ymax = empirical_detection_rate + empirical_detection_rate_mcse)
                  , color = "black") +
    geom_line() +
    geom_point(aes(shape = as.factor(n_max))) +
    facet_wrap(vars(k), nrow = 2, labeller = label_both) +  # nrow = 1 => nebeneinander
    labs(x = "hypothetical number of peeks", y = "empirical false positive rate",
         shape = "upper bound") +
    theme_classic()
```

*Note.* Lineranges show ±1 Monte Carlo standard error (MCSE) of the empirical false-positive rate and are largely smaller than the plotted points; maximum MCSE across all conditions = .003.


```{r}
ggplot(sim_optstop_nmin_summary,
       aes(x = count_peeks_hypothetical, y = empirical_detection_rate)) +
  geom_hline(yintercept = alpha, colour = "grey") +
  geom_linerange(aes(ymin = empirical_detection_rate - empirical_detection_rate_mcse,
                     ymax = empirical_detection_rate + empirical_detection_rate_mcse)
                , color = "black") +
  geom_line(aes(colour = as.factor(k), group = as.factor(k))) +
  geom_point(aes(colour = as.factor(k), shape = as.factor(n_min)), size = 2) +
  labs(x = "hypothetical number of peeks",
       y = "empirical false positive rate",
       colour = "step size",
       shape = "initial n") +
  theme_classic()
```

*Note.* Lineranges show ±1 Monte Carlo standard error (MCSE) of the empirical false-positive rate and are largely smaller than the plotted points; maximum MCSE across all conditions = .003.

### facet plot
```{r}
peeks_cells <- bind_rows(
  sim_optstop_nmax_summary |>
    mutate(dataset = "upper bound varied"), 
  sim_optstop_nmin_summary |>
    mutate(dataset = "initial sample size varied") 
) 

ggplot(peeks_cells,
       aes(x = count_peeks_hypothetical, y = empirical_detection_rate,
           colour = as.factor(k), group = as.factor(k))) +
  geom_hline(yintercept = alpha, colour = "grey") +
  geom_linerange(aes(ymin = empirical_detection_rate - empirical_detection_rate_mcse,
                     ymax = empirical_detection_rate + empirical_detection_rate_mcse)
                , color = "black") +
  geom_line() +
  geom_point(size = 2) +
  facet_wrap(~ dataset) +
  labs(x = "hypothetical number of peeks",
       y = "empirical false positive rate",
       colour = "step size") +
  theme_classic()
```

*Note.* Lineranges show ±1 Monte Carlo standard error (MCSE) of the empirical false-positive rate and are largely smaller than the plotted points; maximum MCSE across all conditions = .003.

## actual peeks in FP results

Note: this chunk only recomputes and saves the summary from the raw data if the cached file does not already exist; otherwise the existing file is loaded.
```{r peeks_at_stop_nmax}
if (!file.exists(optstop_peeks_at_stop_nmax_file)) {
  optstop_peeks_at_stop_nmax <- sim_optstop_nmax |>
    unnest(reported_result) |>
    filter(p_value < alpha) |>
    group_by(count_peeks_hypothetical) |>
    summarize(peeks_median = median(count_peeks_stop))

  saveRDS(optstop_peeks_at_stop_nmax, optstop_peeks_at_stop_nmax_file)
} else {
  optstop_peeks_at_stop_nmax <- readRDS(optstop_peeks_at_stop_nmax_file)
}

optstop_peeks_at_stop_nmax
```


```{r eval=FALSE, include=FALSE}
# TO DO if I want to include have to built n_stop (sample size at stop)
sim_optstop_nmax |>
  unnest(reported_result) |>
  filter(p_value < alpha) |>   
  ggplot(aes(x = n_stop, color = as.factor(k))) +
  stat_ecdf(geom = "step") +
  facet_wrap(~ n_max, scales = "free_x") +
  scale_color_viridis_d(end = 0.9) +
  labs(
    x = "sample size at stopping (n per group)",
    y = "cumulative share of false positives",
    color = "step size"
  ) +
  theme_classic()
```