---
title: "optional stopping results"
author: "Carolin Kirsch"
date: "2026-07-04"
output:
  html_document: default
  word_document: default
  pdf_document:
    latex_engine: xelatex
---

TO DO: sync with MA_quarto.qmd
TO DO: clean up

```{r setup}
library(tidyverse)
library(ggtext)
library(flextable)
library(apa7)
library(ftExtra)

set_flextable_defaults(theme_fun = theme_apa,
                        font.family = "Times New Roman")
```

```{r load_results}
sim_optstop_nmax <- readRDS("results_UBELIX/sim_optstop_nmax_30000iterations.rds")
sim_optstop_nmin <- readRDS("results_UBELIX/sim_optstop_nmin_30000iterations.rds")
```

```{r combine_shared_conditions}
sim_optstop_nmin <- bind_rows(
  sim_optstop_nmin,
  # add the four conditions that shared between the sets and only estimated once
  sim_optstop_nmax |> filter(n_max == 300) 
)
```

```{r general_settings}
# currently both simulation sets run with same fixed alpha and the same number of iterations, if this gets changed later on, adjust
n_iterations <- max(sim_optstop_nmax$iteration) 
alpha <- sim_optstop_nmax$alpha[1] 
```

# replication
## table nmax
Note: this chunk only recomputes and saves the summary if `results_UBELIX/sim_optstop_nmax_summary_<n_iterations>iterations.rds` does not already exist; otherwise the existing file is loaded.
```{r summary_nmax}
sim_optstop_nmax_summary_file <- paste0("results_UBELIX/sim_optstop_nmax_summary_",n_iterations,"iterations.rds")

if (!file.exists(sim_optstop_nmax_summary_file)) {
  sim_optstop_nmax_summary <- sim_optstop_nmax |>
    unnest(reported_result) |>
    group_by(n_max, k, count_peeks_hypothetical) |>
    summarize(
      empirical_detection_rate  = mean(p_value < alpha),
      empirical_detection_rate_mcse = sqrt(empirical_detection_rate * (1 - empirical_detection_rate) / n()),
      .groups = "drop"
      )

  saveRDS(sim_optstop_nmax_summary, sim_optstop_nmax_summary_file)
} else {
  sim_optstop_nmax_summary <- readRDS(sim_optstop_nmax_summary_file)
}

sim_optstop_nmax_summary
```

TO DO: row names, note, caption, etc.
```{r nmax_flextable_table}
sim_optstop_nmax_summary |>
  mutate(`maximum n per condition` = as.character(n_max),
         `step size` = as.character(k)) |>
  mutate(across(where(is.numeric), \(x) scales::number(x, accuracy = 0.001))) |>
  mutate(edr_string = paste0(empirical_detection_rate,
                             " (", 
                             empirical_detection_rate_mcse,
                             ")")) |>
  select(`maximum n per condition`,
         `step size`,
         edr_string) |>
  pivot_wider(names_from = `step size`,
              values_from = edr_string) |>
  column_spanner_label("step size", -1) |>
  apa_flextable(no_format_columns = `maximum n per condition`) |>
  align(j = 1, align = "center") |>
  add_footer_lines(values = as_paragraph(as_i("Note. "), "Values in parentheses are ±1 Monte Carlo Standard Error."))
```
<!-- caption for quarto integration: "Empirical False Positive Rate (±1 MCSE) by Maximum Sample Size and Step Size" -- will be set via #| tbl-cap in the .qmd chunk -->


## table nmin
Note: this chunk only recomputes and saves the summary if `results_UBELIX/sim_optstop_nmin_summary_<n_iterations>iterations.rds` does not already exist; otherwise the existing file is loaded.
```{r summary_nmin}
sim_optstop_nmin_summary_file <- paste0("results_UBELIX/sim_optstop_nmin_summary_",n_iterations,"iterations.rds")

if (!file.exists(sim_optstop_nmin_summary_file)) {
  sim_optstop_nmin_summary <- sim_optstop_nmin |>
    unnest(reported_result) |>
    group_by(n_min, k, count_peeks_hypothetical) |>
    summarize(
      empirical_detection_rate  = mean(p_value < alpha),
      empirical_detection_rate_mcse = sqrt(empirical_detection_rate * (1 - empirical_detection_rate) / n()),
      .groups = "drop"
      )

  saveRDS(sim_optstop_nmin_summary, sim_optstop_nmin_summary_file)
} else {
  sim_optstop_nmin_summary <- readRDS(sim_optstop_nmin_summary_file)
}

sim_optstop_nmin_summary
```

```{r nmin_flextable_table}
sim_optstop_nmin_summary |>
  mutate(`initial n per condition` = as.character(n_min),
         `step size` = as.character(k)) |>
  mutate(across(where(is.numeric), \(x) scales::number(x, accuracy = 0.001))) |>
  mutate(edr_string = paste0(empirical_detection_rate,
                             " (", 
                             empirical_detection_rate_mcse,
                             ")")) |>
  select(`initial n per condition`,
         `step size`,
         edr_string) |>
  pivot_wider(names_from = `step size`,
              values_from = edr_string) |>
  column_spanner_label("step size", -1) |>
  apa_flextable(no_format_columns = `initial n per condition`) |>
  align(j = 1, align = "center") |>
  add_footer_lines(values = as_paragraph(as_i("Note. "), "Values in parentheses are ±1 Monte Carlo Standard Error."))
```
<!-- caption for quarto integration: "Empirical False Positive Rate (±1 MCSE) by Initial Sample Size and Step Size" -- will be set via #| tbl-cap in the .qmd chunk -->

## plot nmax
```{r nmax_plot_repli}
ggplot(sim_optstop_nmax_summary,
       aes(x = k,
           y = empirical_detection_rate,
           color = as.factor(n_max))) +
  geom_hline(yintercept = alpha, col = "grey") +
  geom_linerange(aes(ymin = empirical_detection_rate - empirical_detection_rate_mcse,
                     ymax = empirical_detection_rate + empirical_detection_rate_mcse)
                , color = "black") +
  geom_line() +
  geom_point() +
  geom_text(aes(label = count_peeks_hypothetical),
            nudge_x = 1, vjust = -0.8, size = 3,
            show.legend = FALSE) +
  #scale_y_continuous(breaks = c(0, 0.05, 0.1, 0.2, 0.3, 0.4), limits = c(0,0.45)) +
  scale_x_continuous(breaks = c(1, 5, 10, 50)) +
  labs(
    x = "step size",
    y = "empirical false-positive-rate",
    color = "maximum n per group") +
  theme_classic()
```

## plot nmin
```{r nmin_plot_repli}
ggplot(sim_optstop_nmin_summary,
       aes(x = k,
           y = empirical_detection_rate,
           color = as.factor(n_min))) +
  geom_hline(yintercept = alpha, col = "grey") +
  geom_linerange(aes(ymin = empirical_detection_rate - empirical_detection_rate_mcse,
                     ymax = empirical_detection_rate + empirical_detection_rate_mcse)
                , color = "black") +
  geom_line() +
  geom_point() +
  geom_text(aes(label = count_peeks_hypothetical),
            nudge_x = 1, vjust = -0.8, size = 3,
            show.legend = FALSE) +
  #scale_y_continuous(breaks = c(0, 0.05, 0.1, 0.2, 0.3, 0.4), limits = c(0,0.45)) +
  scale_x_continuous(breaks = c(1, 5, 10, 50)) +
  labs(
    x = "step size",
    y = "empirical false-positive-rate",
    color = "initial n per group") +
  theme_classic()
```

# extension
## span of FPR
```{r nmin_summary_extension}
sim_optstop_nmin_summary |> 
  group_by(k) |>
  summarize(span = max(empirical_detection_rate) - min(empirical_detection_rate))
```

```{r nmax_summary_extension}
sim_optstop_nmax_summary |> 
  group_by(k) |>
  summarize(FPR_span = max(empirical_detection_rate) - min(empirical_detection_rate))
```

## hypothetical peeks
### count hypothetical peeks
```{r}
sim_optstop_nmax |>
  unnest(reported_result) |> 
  group_by(k) |> 
  count(count_peeks_hypothetical, .drop = FALSE) |> 
  mutate(n = n/n_iterations) |> 
  rename(count = n)
```


```{r}
sim_optstop_nmin |>
  unnest(reported_result) |> 
  group_by(k) |> 
  count(count_peeks_hypothetical, .drop = FALSE) |> 
  mutate(n = n/n_iterations) |> 
  rename(count = n)
```

### linear x-axis
```{r}
ggplot(sim_optstop_nmax_summary,
       aes(x = count_peeks_hypothetical, y = empirical_detection_rate)) +
  geom_hline(yintercept = alpha, colour = "grey") +
  geom_linerange(aes(ymin = empirical_detection_rate - empirical_detection_rate_mcse,
                     ymax = empirical_detection_rate + empirical_detection_rate_mcse)
                , color = "black") +
  geom_line(aes(colour = as.factor(k), group = as.factor(k))) +
  geom_point(aes(colour = as.factor(k), shape = as.factor(n_max)), size = 2) +
  labs(x = "hypothetical number of peeks",
       y = "empirical false positive rate",
       colour = "step size",
       shape = "upper bound for n") +
  theme_classic()
```

```{r}
ggplot(sim_optstop_nmax_summary,
       aes(x = count_peeks_hypothetical, y = empirical_detection_rate)) +
    geom_hline(yintercept = alpha, colour = "grey") +
    geom_linerange(aes(ymin = empirical_detection_rate - empirical_detection_rate_mcse,
                       ymax = empirical_detection_rate + empirical_detection_rate_mcse)
                  , color = "black") +
    geom_line() +
    geom_point(aes(shape = as.factor(n_max))) +
    facet_wrap(vars(k), nrow = 2, labeller = label_both) +  # nrow = 1 => nebeneinander
    labs(x = "hypothetical number of peeks", y = "empirical false positive rate",
         shape = "upper bound") +
    theme_classic()
```


```{r}
ggplot(sim_optstop_nmin_summary,
       aes(x = count_peeks_hypothetical, y = empirical_detection_rate)) +
  geom_hline(yintercept = alpha, colour = "grey") +
  geom_linerange(aes(ymin = empirical_detection_rate - empirical_detection_rate_mcse,
                     ymax = empirical_detection_rate + empirical_detection_rate_mcse)
                , color = "black") +
  geom_line(aes(colour = as.factor(k), group = as.factor(k))) +
  geom_point(aes(colour = as.factor(k), shape = as.factor(n_min)), size = 2) +
  labs(x = "hypothetical number of peeks",
       y = "empirical false positive rate",
       colour = "step size",
       shape = "inital n") +
  theme_classic()
```

### facet plot
```{r}
peeks_cells <- bind_rows(
  sim_optstop_nmax_summary |>
    mutate(dataset = "upper bound varied"), 
  sim_optstop_nmin_summary |>
    mutate(dataset = "inital sample size varied") 
) 

ggplot(peeks_cells,
       aes(x = count_peeks_hypothetical, y = empirical_detection_rate,
           colour = as.factor(k), group = as.factor(k))) +
  geom_hline(yintercept = alpha, colour = "grey") +
  geom_linerange(aes(ymin = empirical_detection_rate - empirical_detection_rate_mcse,
                     ymax = empirical_detection_rate + empirical_detection_rate_mcse)
                , color = "black") +
  geom_line() +
  geom_point(size = 2) +
  facet_wrap(~ dataset) +
  labs(x = "hypothetical number of peeks",
       y = "empirical false positive rate",
       colour = "step size") +
  theme_classic()
```

## actual peeks in FP results

```{r}
sim_optstop_nmax |>
  unnest(reported_result) |> 
  filter(p_value < alpha) |> 
  group_by(count_peeks_hypothetical) |> 
  summarize(peeks_median = median(count_peeks_stop))
```


```{r eval=FALSE, include=FALSE}
# TO DO if I want to include have to buit n_stop (sample size at stop)
sim_optstop_nmax |>
  unnest(reported_result) |>
  filter(p_value < alpha) |>   
  ggplot(aes(x = n_stop, color = as.factor(k))) +
  stat_ecdf(geom = "step") +
  facet_wrap(~ n_max, scales = "free_x") +
  scale_color_viridis_d(end = 0.9) +
  labs(
    x = "sample size at stopping (n per group)",
    y = "cumulative share of false positives",
    color = "step size"
  ) +
  theme_classic()
```