---
title: "Outlier exclusion"
author: "Carolin Kirsch"
date: "2026-04-18"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggtext)
library(kableExtra)
library(Dict)
library(furrr)
plan(multisession, workers = parallel::detectCores() - 1)
```

```{r source}
source("functions.R")
```

```{r general_settings}
n_iterations <- 5000
alpha <- 0.05
```

# introduction

Ad-hoc exclusion of extreme values to achieve statistical significance

Number of outlier detection methods: k ∈ {3,5,12}
-> explanation of outlier detection methods in appendix

# functions
## generate data
source generate_data_univar_linreg()

## analyse data
source get_p_univar_linreg()

```{r single_variable_outlier_removed_p_value}
get_p_single_var_outlier_removed <- function(data, out_x, out_y){
  
  has_out_x <- length(out_x) > 0
  has_out_y <- length(out_y) > 0
  
  # if no outliers on any definition: return empty tibble preserving column schema
  # necessary since imap over empty list would return 0x0 tibble
  if (!has_out_x & !has_out_y) return(tibble(type = character(), p_value = double()))
  
  cleaned_data <- list()
  
  if (has_out_x) cleaned_data$out_x <- data |> 
    filter(!(x %in% out_x))
  
  if (has_out_y) cleaned_data$out_y <- data |>
    filter(!(y %in% out_y))
  
  if (has_out_x & has_out_y) cleaned_data$out_xy <- data |>
    filter(!(x %in% out_x & y %in% out_y))
  
  # POSSIBLE EXTENSION small "see everything simulation": To add cleaned_data later: add `cleaned_data = list(cleaned)` inside the tibble() call
  # POSSIBLE EXTENSION more in depth results: add threshold (would need to be extracted in every outlier technique that uses them)
  
  p_tibble <- cleaned_data |>
    # mapping over a named list
    # using its elements (cleaned) and the respective names (type_label)
    imap(\(cleaned, type_label) tibble(
      type    = type_label,
      p_value = get_p_univar_linreg(cleaned )
    )) |>
    bind_rows()
  
  return(p_tibble)
}
```

## outlier detection techniques
### boxplot
define single-variable outliers using graphics::boxplot()
```{r boxplot}
remove_out_boxplot <- function(data) {
  
  # rounding the values is necessary for filtering process later
  # since the boxplot function rounds value to 6 decimal places
  data_rounded <- data |> 
    round(6)
  
  # defining outliers
  out_x <- graphics::boxplot(data_rounded$x, plot = FALSE)$out
  out_y <- graphics::boxplot(data_rounded$y, plot = FALSE)$out

  p_tibble <- data_rounded |> 
    get_p_single_var_outlier_removed(out_x = out_x, out_y = out_y)
  
  return(p_tibble)
}
```

### stem-and-leaf plot
Define single-variable outliers aplpack::stemleaf() function.
```{r stemleaf}
extract_out_stemleaf <- function(var) {
    
    # capture textual output of the stem.leaf function
    utils::capture.output(aplpack::stem.leaf(var)) |>
      
      # extract the lines containing "LO:" or "HI:" (indicate outliers)
      ## this will result in maximum two lines that look like something this: 
      ## "LO: 1.3 1.5 ..." "HI: 5.6 ..."
      keep(\(line) str_detect(line, "LO:") | str_detect(line, "HI:")) |> 
  
      # remove the "LO:" and "HI:", only keep values
      str_remove_all("LO:") |> 
      str_remove_all("HI:") |> 
      
      # remove whitespace and replacing internal whitespace with a single space
      str_squish() |> 
      
      # split so values of outliers are saved individually 
      #(still as character vector)
      str_split(" ") |> 
      
      unlist() |> 
      as.numeric() |> 
      na.omit()
}

remove_out_stemleaf <- function(data) {

  out_x <- extract_out_stemleaf(data$x)
  out_y <- extract_out_stemleaf(data$y)

  p_tibble <- data |> 
    get_p_single_var_outlier_removed(out_x = out_x, out_y = out_y)
  
  return(p_tibble)
}
```

### SD
Stefan & Schönbrodt (2023) define the SD detection technique with subjective threshold setting as follows: "Define single-variable outliers as $x$ standard deviations above or below the mean. Starting with a threshold of $x = 2$ standard deviations, the threshold is increased in increments of 0.5, until no more extreme values exist in the original sample."

```{r SD}
remove_out_sd <- function(data) {
  
  data_z <- data |> 
    mutate(
      x_z = as.numeric(scale(x)),
      y_z = as.numeric(scale(y))
    )
  
  max_x_z <- max(abs(data_z$x_z))
  max_y_z <- max(abs(data_z$y_z))

  # separate threshold sequences per type
  # empty sequence -> no outliers for that type -> no new rows produced  
  thresholds_x  <- if (max_x_z  >= 2) seq(2, max_x_z,  0.5) else numeric(0)
  thresholds_y  <- if (max_y_z  >= 2) seq(2, max_y_z,  0.5) else numeric(0)
  thresholds_xy <- if (min(max_x_z, max_y_z) >= 2) seq(2, min(max_x_z, max_y_z), 0.5) else numeric(0)
  
  # build cleaned datasets per type and deduplicate
  # returns empty datasets in case of no outliers (OPEN check)
  cleaned_x  <- thresholds_x  |> 
    map(\(thr) data_z |> filter(abs(x_z) < thr)) |> 
    unique()
  
  cleaned_y  <- thresholds_y  |>
    map(\(thr) data_z |> filter(abs(y_z) < thr)) |> 
    unique()
  
  cleaned_xy <- thresholds_xy |>
    map(\(thr) data_z |> filter(abs(x_z) < thr & abs(y_z) < thr)) |>
    unique()
  
  # extract p-values
  # in case of empty datasets map preserves schema (OPEN check)
  p_tibble <- bind_rows(
    
    cleaned_x |> 
      map(\(d) tibble(
        type = "out_x",
        p_value = get_p_univar_linreg(d)
        )) |> 
      bind_rows(),
    
    cleaned_y |> 
      map(\(d) tibble(
        type = "out_y",
        p_value = get_p_univar_linreg(d)
        )) |> 
      bind_rows(),
    
    cleaned_xy |> 
      map(\(d) tibble(
        type = "out_xy",
        p_value = get_p_univar_linreg(d)
        )) |>
      bind_rows()
  )
  
  return(p_tibble)
}
```

### percentage
Stefan & Schönbrodt (2023) define the percentage detection technique with subjective threshold setting as follows: "Define single-variable outliers as the highest or lowest $x$ percent of the values. Starting from a threshold of $1/N$ (smallest empirical quantile), the threshold is increased in increments of 0.05 to 1/20, i.e., 5%." 

Since $1/20 = 0.05$, it is illogical to increase the threshold "in increments of 0.05 to 1/20". As I suspected a mistake or typo in the appendix of Stefan & Schönbrodt, I took a quick look at the original code to check the increment definition. Indeed, in their code, the increment was specified as 0.005. I will use this increment in the following code as well. 

```{r percentage}
remove_out_percentage <- function(data) {
  
  n <- nrow(data)
  
  # guard: for potential future application OPEN if I do this here do I have to implement guards everyhwere?
  if(n < 20) stop("too few observations in dataset (<20)")
  
  # thresholds: number of observations to delete from each end
  # rounding number to delete indicated by percentage down conservatively
  n_delete <- 
    seq(from = 1/n, to = 1/20, by = 0.005) |> 
    (\(x) floor(x * n))() |> 
    unique()
  
  # extracing outliers
  ## sort x and y to easily identify extremes
  x_sorted <- sort(data$x)
  y_sorted <- sort(data$y)
  
  ## for each n_delete: identify extreme values on x and y, then get p-values
  p_tibble <- n_delete |>
    map(\(nd) get_p_single_var_outlier_removed(
      data  = data,
      out_x = x_sorted[c(1:nd, (n - nd + 1):n)],
      out_y = y_sorted[c(1:nd, (n - nd + 1):n)]
    )) |>
    bind_rows()
  
  return(p_tibble)
}
```

```{r percentage_add_thresholds? eval=FALSE, include=FALSE}
  thresholds_delete <- tibble(
      percentage = seq(from = 1/n, to = 1/20, by = 0.005),
      n_delete = floor(percentage * n)
    ) |> 
    distinct(n_delete, .keep_all =TRUE)
```

### residuals
```{r residuals_base}
remove_out_residuals_base <- function(data, res) {
  
  data_res <- data |>
    mutate(res_abs = abs( {{ res }} ))
  
  max_res_abs <- max(data_res$res_abs)

  # case 1: no absolute residuals larger than 2 -> remove 3 largest
  if(max_res_abs <= 2) {
    cleaned_data <- data_res |>
      arrange(desc(res_abs)) |>
      slice(-(1:3))
    
    p_tibble <- tibble(
      type = "model",
      p_value = get_p_univar_linreg(cleaned_data)
    )
    
    return(p_tibble)
  }
  
  # case 2: subjective threshold setting from 2 upwards in 0.5 steps
  thresholds <- seq(from = 2, to = max_res_abs, by  = 0.5)
  
  cleaned_data <- thresholds  |> 
    map(\(thr) data_res |> filter(res_abs < thr)) |> 
    unique()
    
  p_tibble <- cleaned_data |> 
    map(\(d) tibble(
        type = "model",
        p_value = get_p_univar_linreg(d)
        )) |>
      bind_rows()
  
  return(p_tibble)
}
```

#### studentized residuals

CK: Stefan & Schönbrodt didn't mention using a package? since they use stats for dfbeta etc. assuming they did the same here (*quicky verified this in the code*)

```{r}
set.seed(1)
data <- generate_data_univar_linreg(n=300)
```

```{r studentized_residuals}
remove_out_stures <- function(data) {
  model <- lm(y ~ x, data)
  remove_out_residuals_base(data, res = stats::rstudent(model))
}
```

#### standardized residuals
Stefan & Schönbrodt (2023) define the studentized residuals detection technique with subjective threshold setting as follows:
Same outlier definition as in (studentized residuals), just with standardized residuals

```{r standardized_residuals}
remove_out_stdres <- function(data) {
  model <- lm(y ~ x, data)
  remove_out_residuals_base(data, res = stats::rstandard(model))
}
```

### DFBETA 
Stefan & Schönbrodt (2023) define the DFBETA detection technique as follows:
Define outliers as values with the three highest DFBETA-influence measures according to stats::dfbeta()

```{r dfbeta}
remove_out_dfbeta <- function(data) {
  model <- lm(y ~ x, data)

  cleaned_data <- data |>
    mutate(dfbeta_x = stats::dfbeta(model)[, 2]) |>
    arrange(desc(abs(dfbeta_x))) |>
    slice(-(1:3))
  
  p_tibble <- tibble(
    type = "model",
    p_value = get_p_univar_linreg(cleaned_data)
  )
  
  return(p_tibble)
}
```

### DFFITS
Stefan & Schönbrodt (2023) define the DFFITS detection technique as follows:
Define outliers as values with absolute DFFIT values larger than $2\sqrt{2/n}$, using the stats::dffits() function.

```{r dffits}
remove_out_dffits <- function(data){
  
  model <- lm(y ~ x, data)
  n <- nrow(data)
  
  cleaned_data <- data |> 
    mutate(dffits_abs = abs(stats::dffits(model))) |>
    filter(dffits_abs <= 2 * sqrt(2/n))
  
  p_tibble <- tibble(
    type = "model",
    p_value = get_p_univar_linreg(cleaned_data)
  )
  
  return(p_tibble)
}
```

### cook's distance
Stefan & Schönbrodt (2023) define the Cook's distance detection technique as follows:
Define values with a Cook's distance larger than the median of an F distribution with p=2 and N-p degrees of freedom, or larger than 1 as outliers, using stats::cooks.distance().

```{r cooks_distance}
remove_out_cook <- function(data) {
  model    <- lm(y ~ x, data)
  n        <- nrow(data)
  f_median <- qf(p = 0.5, df1 = 2, df2 = n - 2)

  cleaned_data <- data |>
    mutate(cooksd = stats::cooks.distance(model)) |>
    filter(cooksd <= 1 & cooksd <= f_median)
  
  p_tibble <- tibble(
    type = "model",
    p_value = get_p_univar_linreg(cleaned_data)
  )

  return(p_tibble)
}
```

### mahalanobis distance
Stefan & Schönbrodt (2023) define the Mahalanobis distance detection technique as follows:
Define outliers as values with a robust Mahalanobis distance larger than $Md^2 > \chi^2(0.98, 2)$ using mvoutlier::uni.plot().
I had to check how the original suppressed the plot bc I couldn't figure it out.

```{r mahalanobis}
remove_out_mahalanobis <- function(data) {
  
  # currently mvoutlier::uniplot() seems unable to handle tibbles
  # PLASTER: transform data to matrix
  data_matrix <- matrix(c(data$x, data$y), ncol=2)
  
  # extract outliers (logical indices) and suppress plot generation
  out <- R.devices::suppressGraphics(mvoutlier::uni.plot(data_matrix))
  
  cleaned_data <- data |>
    mutate(outlier_mahal = out$outliers) |>
    filter(!outlier_mahal)
 
  p_tibble <- tibble(
    type = "model",
    p_value = get_p_univar_linreg(cleaned_data)
  ) 

  return(p_tibble)
}
```

### leverage values
Stefan & Schönbrodt (2023) define the leverage values detection technique as follows:
Define outliers as values with a leverage measure larger than $3 \frac{2}{n}$ using stats::hatvalues().

```{r leverage}
remove_out_leverage <- function(data) {
  model <- lm(y ~ x, data)
  n     <- nrow(data)

  cleaned_data <- data |>
    mutate(leverage = stats::hatvalues(model)) |>
    filter(leverage <= 3 * (2 / n))

  p_tibble <- tibble(
    type = "model",
    p_value = get_p_univar_linreg(cleaned_data)
  ) 
  
  return(p_tibble)
}
```

### covariance ratio
Stefan & Schönbrodt (2023) define the covariance ratio detection technique as follows:
Define outliers with a covariance ration different from 1 as outliers using stats::influence.measures().
CK: but it is never exactly 1?! shouldn't another criteria be used? like this every value is an outlier!!
CK: checked the original code and they used $is.inf[,"cov.r"]) which makes way more sense.

```{r cov_ratio}
remove_out_covratio <- function(data) {
  model <- lm(y ~ x, data)

  cleaned_data <- data |>
    mutate(covr = stats::influence.measures(model)[["is.inf"]][, "cov.r"]) |>
    filter(!covr)
  
  p_tibble <- tibble(
    type = "model",
    p_value = get_p_univar_linreg(cleaned_data)
  ) 

  return(p_tibble)
}

```

## p-hacking
```{r phacking_out_single}
ALL_TECHNIQUES <- c(
  "boxplot", "stemleaf", "SD", 
  "percentage", "student", "standardized",
  "DFBETA", "DFFITS", "cook",
  "mahalanobis", "leverage", "covratio"
)

remove_out_single <- function(data, technique) {
  switch(technique,
    "boxplot" = remove_out_boxplot(data),
    "stemleaf" = remove_out_stemleaf(data),
    "SD" = remove_out_sd(data),
    "percentage"   = remove_out_percentage(data),
    "student"      = remove_out_stures(data),
    "standardized" = remove_out_stdres(data),
    "DFBETA"       = remove_out_dfbeta(data),
    "DFFITS"       = remove_out_dffits(data),
    "cook"         = remove_out_cook(data),
    "mahalanobis"  = remove_out_mahalanobis(data),
    "leverage"     = remove_out_leverage(data),
    "covratio"     = remove_out_covratio(data),
    stop(paste("unknown technique:", technique))
  )
}
```

```{r PLASTER_p_og_single}
get_p_with_original_single <- function(data, technique) {
  bind_rows(
    tibble(
      type = "original", 
      p_value = get_p_univar_linreg(data)
    ),
    remove_out_single(data, technique)
  )
}
```

```{r sequential_removing_and_testing}

sample_techniques <- function(k) {
  sample(ALL_TECHNIQUES, size = k, replace = FALSE)
}

remove_out_seq <- function(data, technique_vec) {
  map(technique_vec, \(tech) remove_out_single(data, tech)) |> 
    bind_rows()
}
```

```{r PLASTER_p_og_seq}
get_p_with_original_seq <- function(data, technique_vec) {
  bind_rows(
    tibble(
      type = "original", 
      p_value = get_p_univar_linreg(data)
    ),
    remove_out_seq(data, technique_vec)
  )
}
```
## reporting strategy
source reporting_strategy_tibbles()

# simulation
POSSIBLE EXTENSION: take the model estimation out of the outlier function and include the model in the grid for better computing efficiency? Would complicate things now and right now it does not seem to be that much of a computational problem

## single
```{r sim_par_out_single, eval=FALSE, include=FALSE}
par_grid_out_single <- expand_grid(
  n                  = c(30, 50, 100, 300),
  iteration          = 1:n_iterations,
  outlier_technique  = ALL_TECHNIQUES,
  reporting_strategy = c("first significant", "smallest", "smallest significant")
)
```

```{r sim_out_single, eval=FALSE, include=FALSE}
set.seed(12345)

sim_out_single <- par_grid_out_single |>
  mutate(
    generated_data = pmap(
      list(n),
      generate_data_univar_linreg
    ),
    # OPEN (possible optimisation): pre-compute model in simulation? and not each time in the technique that need it
    # original_analysis = pmap(
    #   list(generated_data),
    # ),
    analysis_result = future_pmap(   # parallelising
      .l = list(generated_data, outlier_technique),
      .f = get_p_with_original_single,
      .progress = TRUE,
      .options = furrr_options(seed = TRUE)
    ),
    reported_result = pmap(
      list(analysis_result, reporting_strategy),
      reporting_strategy_tibbles
    )
  )
```

## sequentially
```{r sim_par_out_seq}
par_grid_out_seq <- expand_grid(
  n                  = c(30, 50, 100, 300),
  iteration          = 1:n_iterations,
  number_techniques  = c(3, 5, 12),
  reporting_strategy = c("first significant", "smallest", "smallest significant")
)
```

```{r sim_out_seq}
set.seed(12345)

sim_out_seq <- par_grid_out_seq |>
  mutate(
    generated_data = pmap(
      list(n),
      generate_data_univar_linreg
    ),
    # OPEN (possible optimisation): pre-compute model in simulation? and not each time in the technique that need it
    # original_analysis = pmap(
    #   list(generated_data),
    # ),
    techniques_sample = pmap(
      list(number_techniques),
      sample_techniques
    ), 
    analysis_result = future_pmap(   # parallelising
      .l = list(generated_data, techniques_sample),
      .f = get_p_with_original_seq,
      .progress = TRUE,
      .options = furrr_options(seed = TRUE)
    ),
    reported_result = pmap(
      list(analysis_result, reporting_strategy),
      reporting_strategy_tibbles
    )
  )
```

```{r}
saveRDS(sim_out_seq, paste0("results/sim_out_seq_",n_iterations,"iterations.rds"))
```

# results
## table 
```{r}
sim_out_seq_summary <- sim_out_seq |>
  unnest(analysis_result) |> 
  mutate(n = as.factor(n)#,
         #k = as.factor(k)
         ) |> 
  group_by(n) |>
  dplyr::summarize(false_positive_rate = mean(p < alpha), .groups = "drop")

sim_out_seq_summary
```

## plot
```{r}
ggplot(sim_optstop_nmax_summary,
       aes(x = k,
           y = false_positive_rate,
           group = n_max,
           color = n_max)) +
  geom_line() +
  geom_point() +
  labs(
    x = "step size",
    y = "False-Positive-Rate",
    color = "n_max"
  ) +
  theme_minimal()
```


